---
title: "Lab_data_w/o_no-standstill_starts"
author: "Kayla Pehl"
date: "10/1/2020"
output: html_document
---

# This dataset does NOT include trials where the lizard did not start from a standstill

```{r}
#Importing data and required packages

lab.data <- read.csv("~/College of Charleston/Thesis/Videos Digitized/Thesis_videos_digitized_condensed_final_10.15.20_removed_non-standstill_starts+NAs+negatives.csv")

#Creating new data frame without N/A's in bipedal percent column    
##148 rows in lab.data and 88 rows in lab.data_omit, so 60 trials didn't contain bipedalism but 88 did 
lab.data_omit <- lab.data[!is.na(lab.data$bipedal_percent), ]

library(ggplot2)
library(rcompanion)
library(lme4)
library(sjstats)
library(doBy)
library(car)
library(emmeans)
library(multcompView)
library(nlme)
library(lmtest)
```

```{r}
#Checking to see if all data is in correct formats (i.e. factor vs. numeric)    

is.factor(lab.data$ID)
is.factor(lab.data$sex)
is.factor(lab.data$substrate_type)
is.numeric(lab.data$pauses_side)
is.numeric(lab.data$pauses_top)
is.factor(lab.data$bipedal_quad_side)
is.factor(lab.data$bipedal_quad_top)
is.numeric(lab.data$bipedal_side)
is.numeric(lab.data$bipedal_top)
is.numeric(lab.data$total_video_frames)
is.numeric(lab.data$bipedal_percent)
is.factor(lab.data$no_start_from_standstill)
is.factor(lab.data$stops_before_end_of_racetrack)
is.numeric(lab.data$calibration)
is.numeric(lab.data$side_amax)
is.numeric(lab.data$side_pmax)
is.numeric(lab.data$side_posmax)
is.numeric(lab.data$side_spline_tol)
is.numeric(lab.data$side_tmax)
is.numeric(lab.data$side_vmax)
is.numeric(lab.data$top_amax)
is.numeric(lab.data$top_pmax)
is.numeric(lab.data$top_posmax)
is.numeric(lab.data$top_spline_tol)
is.numeric(lab.data$top_tmax)
is.numeric(lab.data$top_vmax)

#All are correct. Therefore nothing in dataset needs to change.
```


#Getting a feel for the data

```{r}
#Summary of data
summary(lab.data)
```

```{r}
#Number of Pauses
table(lab.data$pauses_side)
table(lab.data$pauses_top)
```
```{r}
#Runs per lizard on each substrate type
table(lab.data$ID, lab.data$substrate_type)

#using this data set (without no-standstill starts), lizard F has no runs for D or s, lizard N has no runs for D, and lizard P has no runs for D; all other lizards have at least 1 run for each of the substrates
```
> should this impact the dataset I use? (not using no-standstill starts currently)?
>> Eric said to do data work up using these 3 lizards and then when I'm done I'll need to go back and run stats without the 3 lizards and then compare the results; if the results are the same, we can run the stats including the 3 lizards, if not, we'll have to decide how to proceed


```{r}
#Females vs males count
table(lab.data$sex)

#70% runs are female, 30% are male
```



> Creating histograms to see distribution of data

#Max acceleration
```{r}
#Side max acceleration per substrate 
ggplot(data = lab.data, aes(side_amax))  + 
  geom_histogram(fill="orange", bins = 20, binwidth = 10) +
  facet_wrap(~substrate_type) +
  labs(title = "Side max acceleration per substrate")

#mean, median & range of side_amax data separated into substrate types
summaryBy(side_amax ~ substrate_type, data = lab.data,
          FUN = list(mean, median, min, max))
```

```{r}
#Side max acceleration per sex
ggplot(data = lab.data, aes(side_amax))  + 
  geom_histogram(fill="orange", bins = 20, binwidth = 5) +
  facet_wrap(~sex) +
  labs(title = "Side max acceleration per sex")

#mean, median & range of side max acceleration separated into sex
summaryBy(side_amax ~ sex, data = lab.data,
          FUN = list(mean, median, min, max))
```

```{r}
#Top max acceleration per substrate 
ggplot(data = lab.data, aes(top_amax))  + 
  geom_histogram(fill="purple", bins = 30, binwidth = 10) +
  facet_wrap(~substrate_type) +
  labs(title = "Top max acceleration per substrate")

#mean, median & range of top max acceleration separated into substrate types
summaryBy(top_amax ~ substrate_type, data = lab.data,
          FUN = list(mean, median, min, max))
```

```{r}
#Top max acceleration per sex
ggplot(data = lab.data, aes(top_amax))  + 
  geom_histogram(fill="purple", bins = 30, binwidth = 5) +
  facet_wrap(~sex) +
  labs(title = "Top max acceleration per sex")

#mean, median & range of top max acceleration separated into sex
summaryBy(top_amax ~ sex, data = lab.data,
          FUN = list(mean, median, min, max))
```


#Max velocity
```{r}
#Side max velocity per substrate 
ggplot(data = lab.data, aes(side_vmax))  + 
  geom_histogram(fill="orange", bins = 10) +
  facet_wrap(~substrate_type) +
  labs(title = "Side max velocity per substrate")

#mean, median & range of side_vmax data separated into substrate types
summaryBy(side_vmax ~ substrate_type, data = lab.data,
          FUN = list(mean, median, min, max))
```

```{r}
#Side max velocity per sex
ggplot(data = lab.data, aes(side_vmax))  + 
  geom_histogram(fill="orange", bins = 10) +
  facet_wrap(~sex) +
  labs(title = "Side max velocity per sex")

#mean, median & range of side vmax velocity separated into sex
summaryBy(side_vmax ~ sex, data = lab.data,
          FUN = list(mean, median, min, max))
```

```{r}
#Top max velocity per substrate 
ggplot(data = lab.data, aes(top_vmax))  + 
  geom_histogram(fill="purple", bins = 10) +
  facet_wrap(~substrate_type) +
  labs(title = "Top max velocity per substrate")

#mean, median & range of top max velocity separated into substrate types
summaryBy(top_vmax ~ substrate_type, data = lab.data,
          FUN = list(mean, median, min, max))
```

```{r}
#Top max velocity per sex
ggplot(data = lab.data, aes(top_vmax))  + 
  geom_histogram(fill="purple", bins = 10) +
  facet_wrap(~sex) +
  labs(title = "Top max velocity per sex")

#mean, median & range of top max velocity separated into sex
summaryBy(top_vmax ~ sex, data = lab.data,
          FUN = list(mean, median, min, max))
```



#Bipedalism Percentage
```{r}
#Bipedalism percentage per sex Histogram
ggplot(data = lab.data_omit, aes(as.numeric(bipedal_percent)))  + 
  geom_histogram(fill="dark green", bins = 20, binwidth = 10) +
  facet_wrap(~sex)

#mean, median & range of bipedal percentage separated into sex
summaryBy(bipedal_percent ~ sex, data = lab.data_omit,
          FUN = list(mean, median, min, max))

#males are bipedal a lot more than females
```

```{r}
#Bipedalism percentage per substrate Histogram
ggplot(data = lab.data_omit, aes(as.numeric(bipedal_percent)))  + 
  geom_histogram(fill="dark green", bins = 20, binwidth = 10) +
  facet_wrap(~substrate_type)

summaryBy(bipedal_percent ~ substrate_type, data = lab.data_omit,
          FUN = list(mean, median, min, max))

#lizards are a lot more likely to run bipedal on debris than sand
```


```{r}
#Bipedalism percentage per substrate and sex Histogram
ggplot(data = lab.data_omit, aes(as.numeric(bipedal_percent)))  + 
  geom_histogram(fill="dark green", bins = 20, binwidth = 10) +
  facet_wrap(~substrate_type + sex) +
  facet_grid(sex ~ substrate_type)

#mean, median & range of bipedal percentage separated into substrate type
summaryBy(bipedal_percent ~ substrate_type + sex, data = lab.data_omit,
          FUN = list(mean, median, min, max))

#males are bipedal more often on every substrate than females

##for final manuscript, change plot titles to actual names instead of letters
### https://www.zevross.com/blog/2019/04/02/easy-multi-panel-plots-in-r-using-facet_wrap-and-facet_grid-from-ggplot2/ 
```


```{r}
#Boxplots of Bipedalism on acceleration 

ggplot(data = lab.data, aes(x=bipedal_quad_side, y=side_amax)) +
  geom_boxplot() +
    labs(title = "Bipedalism effect on max acceleration (side)")

ggplot(data = lab.data, aes(x=bipedal_quad_top, y=top_amax)) +
  geom_boxplot() +
    labs(title = "Bipedalism effect on max acceleration (top)")
```

```{r}
#Boxplots of Bipedalism on velocity

ggplot(data = lab.data, aes(x=bipedal_quad_side, y=side_vmax)) +
  geom_boxplot() +
    labs(title = "Bipedalism effect on max velocity (side)")

ggplot(data = lab.data, aes(x=bipedal_quad_top, y=top_vmax)) +
  geom_boxplot() +
    labs(title = "Bipedalism effect on max velocity (top)")
```


#Pmax, Posmax, Tmax
```{r}
#Histograms of pmax, posmax and tmax
ggplot(data = lab.data, aes(side_pmax)) +
  geom_histogram(fill = "dark blue") +
    labs(title = "Side Pmax")                     #skewed right

ggplot(data = lab.data, aes(side_posmax)) +
  geom_histogram(fill = "dark blue") +
    labs(title = "Side Posmax")                   #skewed left

ggplot(data = lab.data, aes(side_tmax)) +
  geom_histogram(fill = "dark blue") +
    labs(title = "Side Tmax")

ggplot(data = lab.data, aes(top_pmax)) +
  geom_histogram(fill = "medium blue") +
    labs(title = "Top Pmax")                      #skewed right

ggplot(data = lab.data, aes(top_posmax)) +
  geom_histogram(fill = "medium blue") +
    labs(title = "Top Posmax")

ggplot(data = lab.data, aes(top_tmax)) +
  geom_histogram(fill = "medium blue") +
    labs(title = "Top Tmax")                      #skewed right
```

```{r}
#Looking at bipedal vs. quadrupedal percentages from side and top

table(lab.data$bipedal_quad_side)
#42.6% bipedal on side runs, 57.4% quadrupedal on side runs

table(lab.data$bipedal_quad_top)
#53.3% bipedal on top runs, 46.7% quadrupedal on top runs
```


# Checking for normality

```{r}
#Checking for normality of data: max accelerations and velocities

#histograms
ggplot(data = lab.data, aes(side_amax))  + 
  geom_histogram(bins = 11, col = ("grey")) +
  labs(title = "Side Amax")                        #looks pretty normal

ggplot(data = lab.data, aes(top_amax))  + 
  geom_histogram(bins = 11, col = ("grey")) +
  labs(title = "Top Amax")                         #looks skewed right

ggplot(data = lab.data, aes(side_vmax))  + 
  geom_histogram(bins = 13, col = ("grey")) +
  labs(title = "Side Vmax")                        #looks fairly normal

ggplot(data = lab.data, aes(top_vmax))  + 
  geom_histogram(bins = 11, col = ("grey")) +
  labs(title = "Top Vmax")                         #looks skewed left

#Shapiro-Wilks normality test (p-value >0.05 assumes normality)
shapiro.test(lab.data$side_amax)
shapiro.test((lab.data$top_amax))
shapiro.test((lab.data$side_vmax))
shapiro.test(lab.data$top_vmax)

#side_amax: not normal
#top_amax: not normal
#side_vmax: normal
#top_vmax: normal
```

```{r}
#Checking for normality of data: time and position

#histograms
ggplot(data = lab.data, aes(side_tmax))  + 
  geom_histogram(bins = 11, col = ("grey")) +
  labs(title = "Side Tmax")                           #looks skewed right

ggplot(data = lab.data, aes(top_tmax))  + 
  geom_histogram(bins = 11, col = ("grey")) +
    labs(title = "Top Tmax")                          #looks skewed right

ggplot(data = lab.data, aes(side_posmax))  + 
  geom_histogram(bins = 13, col = ("grey")) +
    labs(title = "Side Posmax")                       #looks fairly normal

ggplot(data = lab.data, aes(top_posmax))  + 
  geom_histogram(bins = 30, col = ("grey")) +
    labs(title = "Top Posmax")

#Shapiro-Wilks normality test (p-value >0.05 assumes normality)
shapiro.test(lab.data$side_tmax)
shapiro.test((lab.data$top_tmax))
shapiro.test((lab.data$side_posmax))
shapiro.test(lab.data$top_posmax)

#side_tmax: not normal
#top_tmax: not normal
#side_posmax: normal
#top_posmax: not normal
```

```{r}
#Checking for normality of data: bipedal_side and bipedal_top

#histograms
ggplot(data = lab.data_omit, aes(bipedal_side))  + 
  geom_histogram(bins = 20, col = ("grey"))

ggplot(data = lab.data_omit, aes(bipedal_top))  + 
  geom_histogram(bins = 20, col = ("grey"))

#Shapiro-Wilks normality test (p-value >0.05 assumes normality)
shapiro.test(lab.data_omit$bipedal_side)
shapiro.test((lab.data_omit$bipedal_top))

#bipedal_side: normal
#bipedal_top: not normal
```



> Looking at diagnostics data

```{r}
#Top max acceleration vs. bipedalism in top

top.amax.biped <- lm(top_amax ~ bipedal_top, data = lab.data)
res.top.amax.biped <- resid(top.amax.biped)
plot(fitted(top.amax.biped), res.top.amax.biped)
abline(0,0)

qqnorm(res.top.amax.biped)
qqline(res.top.amax.biped,
       col = 'red')
plot(density(res.top.amax.biped))

plot(top.amax.biped)

#QQ plot shows data is skewed left -- need to transform
```

```{r}
#Side camera max acceleration vs. bipedalism in side
side.amax.biped <- lm(side_amax ~ bipedal_side, data = lab.data)
res.side.amax.biped <- resid(side.amax.biped)
plot(fitted(side.amax.biped), res.side.amax.biped)

qqnorm(res.side.amax.biped)
qqline(res.side.amax.biped,
       col = 'red')
plot(density(res.side.amax.biped))

plot(side.amax.biped)

#QQ plot maybe slightly skewed left - need to transform
```

```{r}
#Top max velocity vs. bipedalism in top
top.vmax.biped <- lm(top_vmax ~ bipedal_top, data = lab.data)
res.top.vmax.biped <- resid(top.vmax.biped)
plot(fitted(top.vmax.biped), res.top.vmax.biped)

qqnorm(res.top.vmax.biped)
qqline(res.top.vmax.biped,
       col = 'red')
plot(density(res.top.vmax.biped))

plot(top.vmax.biped)

#QQ plot maybe slightly skewed left but it looks pretty close to normal
```

```{r}
#Side max velocity vs. bipedalism in side
side.vmax.biped <- lm(side_vmax ~ bipedal_side, data = lab.data)
res.side.vmax.biped <- resid(side.vmax.biped)
plot(fitted(side.vmax.biped), res.side.vmax.biped)

qqnorm(res.side.vmax.biped)
qqline(res.side.vmax.biped,
       col = 'red')
plot(density(res.side.vmax.biped))

plot(side.vmax.biped)

#QQ plot looks pretty close to normal 
```


#Trying to determine if v_max needs to be transformed or not

```{r}
#Log transform side max velocity vs. bipedalism in side

log_side_vmax <- log(lab.data$side_vmax)
log.side.vmax.biped <- lm(log_side_vmax ~ bipedal_side, data = lab.data)
res.log.side.vmax.biped <- resid(log.side.vmax.biped)
plot(fitted(log.side.vmax.biped), res.log.side.vmax.biped)

shapiro.test(res.log.side.vmax.biped)
hist(log_side_vmax)

qqnorm(res.log.side.vmax.biped)
qqline(res.log.side.vmax.biped,
       col = 'red')
plot(density(res.log.side.vmax.biped))  #QQ plot looks pretty close to normal

plot(log.side.vmax.biped)

ggplot(data = lab.data, aes(log_side_vmax))  + 
  geom_histogram(bins = 13, col = ("grey"))    #skewed left

plot(lm(side_vmax ~ bipedal_side, data = lab.data))  #Side vmax vs. bipedalism

#Q-Q plots of regular and log-transformed side vmax data look basically the same to me. I'm not sure if log-transforming did a lot for the data.
```


```{r}
# Added this section during meeting with Eric

#log transform each variable!
#probably need to log transform all variables and then check residuals
#levene's test

log_side_amax <- log10(lab.data$side_amax+1)
log.side.amax.biped <- lm(log_side_amax ~ bipedal_side, data = lab.data)
res.log.side.amax.biped <- resid(log.side.amax.biped)
plot(fitted(log.side.amax.biped), res.log.side.amax.biped)


shapiro.test(res.log.side.amax.biped)
hist(res.log.side.amax.biped)

leveneTest(log_side_amax, lab.data$bipedal_side)
```

```{r}
#Comparing side vmax data from original with log-transformed

#histograms
plotNormalHistogram(lab.data$side_vmax)  #appears pretty symmetrical 
plotNormalHistogram(log_side_vmax)  #skew left

#Shapiro-Wilks test for normality
shapiro.test(lab.data$side_vmax)  #normal -- not transformed
shapiro.test(log_side_vmax)  #not normal -- log transformed
```

```{r}
#Log transform top max velocity vs. bipedalism in top

log_top_vmax <- log(lab.data$top_vmax)
log.top.vmax.biped <- lm(log_top_vmax ~ bipedal_top, data = lab.data)
res.log.top.vmax.biped <- resid(log.top.vmax.biped)
plot(fitted(log.top.vmax.biped), res.log.top.vmax.biped)  

qqnorm(res.log.top.vmax.biped)
qqline(res.log.top.vmax.biped,
       col = 'red')
plot(density(res.log.top.vmax.biped))  #QQ plot looks pretty close to normal

plot(log.top.vmax.biped)

plot(lm(top_vmax ~ bipedal_top, data = lab.data))  #not transformed  #QQ plot is skewed left
```

```{r}
#Comparing top vmax data from original with log-transformed

#histograms
plotNormalHistogram(lab.data$top_vmax)  #appears pretty symmetrical 
plotNormalHistogram(log_top_vmax)  #skew left

#Shapiro-Wilks test for normality
shapiro.test(lab.data$top_vmax)  #normal -- no transformation
shapiro.test(log_top_vmax)  #not normal  -- log transformed
```



# Linear Models with Interaction Effects

```{r}
#Velocity

anova(lm(top_vmax ~ sex*bipedal_top*substrate_type*ID, data=lab.data))
#sex, bipedal_top & substrate type are all significant by themselves

anova(lm(side_vmax ~ sex*bipedal_side*substrate_type*ID, data = lab.data))
#sex and bipedalism were very significant by themselves
#ID was also significant by itself
```

```{r}
#Acceleration

anova(lm(top_amax ~ sex*bipedal_top*substrate_type*ID, data = lab.data))
#nothing is significant

anova(lm(side_amax ~ sex*bipedal_side*substrate_type*ID, data = lab.data))
#nothing is significant
```


# Bipedal Percentage

```{r}
#Bipedal Percentage Plots

#Histogram of bipedal percent without N/A's broken into sex
ggplot(data = lab.data_omit, aes(as.numeric(bipedal_percent), col=sex, fill=sex)) +
  geom_histogram(col = "white") + 
  labs(title = "Histogram of Bipedal Percent (without N/As) Broken Down by Sex")

#Histogram of bipedal percent without N/A's separated into substrate type
ggplot(data = lab.data_omit, aes(as.numeric(bipedal_percent), col=substrate_type, fill=substrate_type)) +
  geom_histogram(bins = 20, binwidth = 8) +
  facet_wrap(~substrate_type) +
  labs(title = "Histogram of Bipedal Percent (without N/As) Separated by Substrate Type")

#Histogram of bipedal percent without NA's separated into substrate type and sex
ggplot(data = lab.data_omit, aes(as.numeric(bipedal_percent), col=substrate_type, fill=substrate_type, alpha=sex)) +
  geom_histogram(bins = 20, binwidth = 8) +
  facet_wrap(~substrate_type) +
  labs(title = "Histogram of Bipedal Percent (without N/As) Separated by Substrate Type and Sex")

#Histogram of bipedal percent without N/A's broken into substrate type
ggplot(data = lab.data_omit, aes(as.numeric(bipedal_percent), col=substrate_type, fill=substrate_type)) +
  geom_histogram(bins = 20, binwidth = 8) +
  labs(title = "Histogram of Bipedal Percent (without N/As) Broken into Substrate Type")

#Boxplot of bipedal percent without N/A's separated by substrate type
ggplot(data = lab.data_omit, aes(substrate_type, as.numeric(bipedal_percent))) +
  geom_boxplot() + 
  labs(title = "Boxplot of Bipedal Percent (without N/As) by Substrate Type")

#Boxplot of bipedal percent without N/A's separated by sex
ggplot(data = lab.data_omit, aes(sex, as.numeric(bipedal_percent))) +
  geom_boxplot() + 
  labs(title = "Boxplot of Bipedal Percent (without N/As) by Sex")

#Boxplot of bipedal percent without N/A's separated by substrate type and sex
ggplot(data = lab.data_omit, aes(substrate_type, as.numeric(bipedal_percent), fill = sex)) +
  geom_boxplot() +
  labs(title = "Boxplot of Bipedal Percent (without N/As) by Substrate Type and Sex")


#Lizards are running bipedal the most on debris compared to the other substrate types.
#Based on the boxplot, lizards are treating bipedalism on sand/debris as more similar to sand than debris
#Males run bipedally more frequently than females for all substrate types
```


```{r}
#Percent bipedal vs. substrate types and sex linear models
summary(lm(as.numeric(bipedal_percent) ~ substrate_type*sex, data = lab.data_omit))
anova((lm(as.numeric(bipedal_percent) ~ substrate_type*sex, data = lab.data_omit)))

#Neither substrate type nor sex is significantly effecting bipedalism according to the models
```

```{r}
#looked at with Eric during meeting 10/22

#Mixed model bipedal percent and substrate type
lmer2<-lmer(bipedal_percent ~ substrate_type + (1 | ID), data = lab.data_omit)
lmer3<- lmer(sqrt(bipedal_percent) ~ substrate_type + (1 | ID), data = lab.data_omit)
summary(lmer(bipedal_percent ~ substrate_type + (1 | ID), data = lab.data_omit))
Anova(lmer2)

emmeans(lmer2, pairwise ~ substrate_type, adjust = "tukey", data=lab.data)

hist(residuals(lmer2))
hist(residuals(lmer3))

shapiro.test(residuals(lmer2))  #not normal
shapiro.test(residuals(lmer3))  #need sqrt transform - fixed normality issue   #normal

#sand and sand/debris have significant p-values
```

```{r}
lmer.biped.sub <- lmer(sqrt(bipedal_percent) ~ substrate_type + (1|ID), data = lab.data_omit)
anova(lmer.biped.sub)
plotNormalHistogram(residuals(lmer.biped.sub))   #appears normal
lmer.biped.sub.tukey <- emmeans(lmer.biped.sub, pairwise ~ substrate_type, adjust="tukey", data=lab.data)   
lmer.biped.sub.tukey   #significant differences between D-S but not for D-SD and S-SD
ggplot(data = lab.data_omit, aes(x= substrate_type, y=bipedal_percent)) +
  geom_boxplot() +
  labs(title = "Bipedal Percent by Substrate Type") 
plot(lmer.biped.sub.tukey)
summary(lmer.biped.sub)
performance::r2(lmer.biped.sub)  #low R2s

#can't figure out how to do a Tukey test for lme function so had to use lmer instead
```


```{r}
#Mixed model bipedal percent and sex
lmer(bipedal_percent ~ sex + (1 | ID), data = lab.data_omit)
summary(lmer(bipedal_percent ~ sex + (1 | ID), data = lab.data_omit))
```

```{r}
#Mixed model bipedal percent and substrate type + sex
lmer1<-lmer(bipedal_percent ~ sex*substrate_type + (1 | ID), data = lab.data_omit)
summary(lmer(bipedal_percent ~ sex*substrate_type + (1 | ID), data = lab.data_omit))

#no interaction effects are significant
```

> Levene's test

## Ho: the groups we're comparing all have equal population variances. If this is true, we'll probably find slightly different variances in our samples from these populations. If the sample variances have very different variances then it'll suggest that the population variances weren't equal after all and Ho would have to be rejected.
## Ha: the variance among the groups is not equal
## 2 assumptions: independent observations and the test variable is quantitative
## if p-value is less than significance level (0.05), we can reject Ho and conclude that the variance among the groups is NOT equal
## we want them to not be significant


```{r}
#Levene test - ID
leveneTest(side_amax ~ ID, data = lab.data)   #not significant
leveneTest(top_amax ~ ID, data = lab.data)   #not significant

leveneTest(side_vmax ~ ID, data = lab.data)   #not significant
leveneTest(top_vmax ~ ID, data = lab.data)   #not significant
```

```{r}
#Levene test - sex
leveneTest(side_amax ~ sex, data = lab.data)   #not significant
leveneTest(top_amax ~ sex, data = lab.data)   #not significant

leveneTest(side_vmax ~ sex, data = lab.data)   #not significant
leveneTest(top_vmax ~ sex, data = lab.data)   #not significant
```

```{r}
#Levene test - bipedalism
leveneTest(bipedal_percent ~ substrate_type, data = lab.data)  #not significant so can't reject Ho; variance among groups may be equal
leveneTest(bipedal_side ~ substrate_type, data = lab.data)   #not significant
leveneTest(bipedal_top ~ substrate_type, data = lab.data)   #not significant
```




> Checking assumptions: normality of residuals
## all residuals should be log transformed

```{r}
#Log transforming all variables 
##included log10 and +1 because it forces 0's to be 0's and not something else

log_side_amax <- log10(lab.data$side_amax +1)
log_top_amax <- log10(lab.data$top_amax +1)
log_side_vmax <- log10(lab.data$side_vmax +1)
log_top_vmax <- log10(lab.data$top_vmax +1)
log_side_posmax <- log10(lab.data$side_posmax +1)
log_top_posmax <- log10(lab.data$top_posmax +1)
log_side_tmax <- log10(lab.data$side_tmax +1)
log_top_tmax <- log10(lab.data$top_tmax +1)
log_side_pmax <- log10(lab.data$side_pmax +1)
log_top_pmax <- log10(lab.data$top_pmax +1)

log_bipedal_side <- log10(lab.data$bipedal_side +1)
log_bipedal_top <- log10(lab.data$bipedal_top +1)
log_bipedal_percent <- log10(lab.data$bipedal_percent +1)
```


#Writing linear models and Residuals

```{r}
#log.side.amax.biped
log.side.amax.biped <- lm(log_side_amax ~ log_bipedal_side, data = lab.data)
res.log.side.amax.biped <- resid(log.side.amax.biped)
plot(fitted(log.side.amax.biped), res.log.side.amax.biped)

qqnorm(res.log.side.amax.biped)
qqline(res.log.side.amax.biped,
       col = 'red')
shapiro.test(res.log.side.amax.biped)  #normal
plotNormalHistogram(res.log.side.amax.biped)

#log transforming bipedal_side did not appear to change ANYTHING, so I'm just going to run all other runs using log_bipedal

#lmer model
lmer.log.side.amax.biped <- lmer(log_side_amax ~ log_bipedal_side + substrate_type + sex + (1 | ID), data = lab.data)
anova(lmer.log.side.amax.biped)
```


```{r} 
#log.top.amax.biped
log.top.amax.biped <- lm(log_top_amax ~ log_bipedal_top, data = lab.data)
res.log.top.amax.biped <- resid(log.top.amax.biped)

plot(fitted(log.top.amax.biped), res.log.top.amax.biped)
qqnorm(res.log.top.amax.biped)
qqline(res.log.top.amax.biped,
       col = 'red')
shapiro.test(res.log.top.amax.biped)  #normal
plotNormalHistogram(res.log.top.amax.biped)

#lmer model
lmer.log.top.amax.biped <- lmer(log_top_amax ~ log_bipedal_top + substrate_type + sex + (1 | ID), data = lab.data)
anova(lmer.log.top.amax.biped)
```

```{r}
#log.side.vmax.biped
log.side.vmax.biped <- lm(log_side_vmax ~ bipedal_side, data = lab.data)
res.log.side.vmax.biped <- resid(log.side.vmax.biped)

plot(fitted(log.side.vmax.biped), res.log.side.vmax.biped)
qqnorm(res.log.side.vmax.biped)
qqline(res.log.side.vmax.biped,
       col = 'red')
shapiro.test(res.log.side.vmax.biped)  #normal
plotNormalHistogram(res.log.side.vmax.biped)

#lmer model
lmer.log.side.vmax.biped <- lmer(log_side_vmax ~ bipedal_side + substrate_type + sex + (1 | ID), data = lab.data)
anova(lmer.log.side.vmax.biped)
```

```{r}
#log.top.vmax.biped
log.top.vmax.biped <- lm(log_top_vmax ~ bipedal_top, data = lab.data)
res.log.top.vmax.biped <- resid(log.top.vmax.biped)

plot(fitted(log.top.vmax.biped), res.log.top.vmax.biped)
qqnorm(res.log.top.vmax.biped)
qqline(res.log.top.vmax.biped,
       col = 'red')
shapiro.test(res.log.top.vmax.biped)  #normal
plotNormalHistogram(res.log.top.vmax.biped)

#lmer model
lmer.log.top.vmax.biped <- lmer(log_top_vmax ~ bipedal_top + substrate_type + sex + (1 | ID), data = lab.data)
anova(lmer.log.top.vmax.biped)
```

```{r}
#log.side.posmax.biped
log.side.posmax.biped <- lm(log_side_posmax ~ bipedal_side, data = lab.data)
res.log.side.posmax.biped <- resid(log.side.posmax.biped)

plot(fitted(log.side.posmax.biped), res.log.side.posmax.biped)
qqnorm(res.log.side.posmax.biped)
qqline(res.log.side.posmax.biped,
       col = 'red')
shapiro.test(res.log.side.posmax.biped)  #normal
plotNormalHistogram(res.log.side.posmax.biped)

#lmer model
lmer.log.side.posmax.biped <- lmer(log_side_posmax ~ bipedal_side + substrate_type + sex + (1 | ID), data = lab.data)
anova(lmer.log.side.posmax.biped)
```

```{r}
#log.top.posmax.biped
log.top.posmax.biped <- lm(log_top_posmax ~ bipedal_top, data = lab.data)
res.log.top.posmax.biped <- resid(log.top.posmax.biped)

plot(fitted(log.top.posmax.biped), res.log.top.posmax.biped)
qqnorm(res.log.top.posmax.biped)
qqline(res.log.top.posmax.biped,
       col = 'red')
shapiro.test(res.log.top.posmax.biped)  #not normal - skewed right   #really weird data
plotNormalHistogram(res.log.top.posmax.biped)

#lmer model
lmer.log.top.posmax.biped <- lmer(log_top_posmax ~ bipedal_top + substrate_type + sex + (1 | ID), data = lab.data)
anova(lmer.log.top.posmax.biped)
```

```{r}
#log.side.tmax.biped
log.side.tmax.biped <- lm(log_side_tmax ~ bipedal_side, data = lab.data)
res.log.side.tmax.biped <- resid(log.side.tmax.biped)

plot(fitted(log.side.tmax.biped), res.log.side.tmax.biped)
qqnorm(res.log.side.tmax.biped)
qqline(res.log.side.tmax.biped,
       col = 'red')
shapiro.test(res.log.side.tmax.biped)  #not normal - skewed right
plotNormalHistogram(res.log.side.tmax.biped)

#lmer model
lmer.log.side.tmax.biped <- lmer(log_side_tmax ~ bipedal_side + substrate_type + sex + (1 | ID), data = lab.data)
anova(lmer.log.side.tmax.biped)
```

```{r}
#log.top.tmax.biped
log.top.tmax.biped <- lm(log_top_tmax ~ bipedal_top, data = lab.data)
res.log.top.tmax.biped <- resid(log.top.tmax.biped)

plot(fitted(log.top.tmax.biped), res.log.top.tmax.biped)
qqnorm(res.log.top.tmax.biped)
qqline(res.log.top.tmax.biped,
       col = 'red')
shapiro.test(res.log.top.tmax.biped)  #not normal - skewed right
plotNormalHistogram(res.log.top.tmax.biped)

#lmer model
lmer.log.top.tmax.biped <- lmer(log_top_tmax ~ bipedal_top + substrate_type + sex + (1 | ID), data = lab.data)
anova(lmer.log.top.tmax.biped)
```

```{r}
#log.side.pmax.biped
log.side.pmax.biped <- lm(log_side_pmax ~ bipedal_side, data = lab.data)
res.log.side.pmax.biped <- resid(log.side.pmax.biped)

plot(fitted(log.side.pmax.biped), res.log.side.pmax.biped)
qqnorm(res.log.side.pmax.biped)
qqline(res.log.side.pmax.biped,
       col = 'red')
shapiro.test(res.log.side.pmax.biped)  #normal 
plotNormalHistogram(res.log.side.pmax.biped)

#lmer model
lmer.log.side.pmax.biped <- lmer(log_side_pmax ~ bipedal_side + substrate_type + sex + (1 | ID), data = lab.data)
anova(lmer.log.side.pmax.biped)
```

```{r}
#log.top.pmax.biped
log.top.pmax.biped <- lm(log_top_pmax ~ bipedal_top, data = lab.data)
res.log.top.pmax.biped <- resid(log.top.pmax.biped)

plot(fitted(log.top.pmax.biped), res.log.top.pmax.biped)
qqnorm(res.log.top.pmax.biped)
qqline(res.log.top.pmax.biped,
       col = 'red')
shapiro.test(res.log.top.pmax.biped)  #normal
plotNormalHistogram(res.log.top.pmax.biped)

#lmer model
lmer.log.top.pmax.biped <- lmer(log_top_pmax ~ bipedal_top + substrate_type + sex + (1 | ID), data = lab.data)
anova(lmer.log.top.pmax.biped)
```



> Glmer and lmer

## lmer is used to fit linear mixed-effect models, so it assumes that the residual error has a Gaussian distribution. If your dependent variable A is a binary outcome (e.g. a yes/no response), then the error distribution is binomial and not Gaussian. In this case you have to use glmer, which allow to fit a generalized linear mixed-effects model: these models include a link function that allows to predict response variables with non-Gaussian distributions.
## glmer -> use when dependent variable (X) is a binary outcome (e.g. yes/no)
### 1 refers to an intercept; means 'add a random intercept for each ID level'


#Lmer models with v/a/biped/p/t vs. substrate type -- Question 2

#larger F-values represent greater dispersion of data

```{r}
#Velocity
lmer.log.side.vmax.sub <- lmer(log_side_vmax ~ substrate_type + (1 | ID), data = lab.data)
anova(lmer.log.side.vmax.sub)
summary(lmer.log.side.vmax.sub)

lmer.log.top.vmax.sub <- lmer(log_top_vmax ~ substrate_type + (1 | ID), data = lab.data)
anova(lmer.log.top.vmax.sub)  
```

```{r}
#Acceleration
lmer.log.side.amax.sub <- lmer(log_side_vmax ~ substrate_type + (1 | ID), data = lab.data)
anova(lmer.log.side.amax.sub)  

lmer.log.top.amax.sub <- lmer(log_top_vmax ~ substrate_type + (1 | ID), data = lab.data)
anova(lmer.log.top.amax.sub)  
```  

```{r}
#Bipedalism
lmer.log.side.bipedal.sub <- lmer(log_bipedal_side ~ substrate_type + (1 | ID), data = lab.data)
anova(lmer.log.side.bipedal.sub)  

lmer.log.top.bipedal.sub <- lmer(log_bipedal_top ~ substrate_type + (1 | ID), data = lab.data)
anova(lmer.log.top.bipedal.sub)  
```

```{r}
#Power
lmer.log.side.pmax.sub <- lmer(log_side_pmax ~ substrate_type + (1 | ID), data = lab.data)
anova(lmer.log.side.pmax.sub)   

lmer.log.top.pmax.sub <- lmer(log_top_pmax ~ substrate_type + (1 | ID), data = lab.data)
anova(lmer.log.top.pmax.sub)   
```

```{r}
#Time
lmer.log.side.tmax.sub <- lmer(log_side_tmax ~ substrate_type + (1 | ID), data = lab.data)
anova(lmer.log.side.tmax.sub) 

lmer.log.top.tmax.sub <- lmer(log_top_pmax ~ substrate_type + (1 | ID), data = lab.data)
anova(lmer.log.top.tmax.sub)  
```


```{r}
#Velocity - sex
lmer.log.side.vmax.sex <- lmer(log_side_vmax ~ sex + (1 | ID), data = lab.data)
anova(lmer.log.side.vmax.sex)  

lmer.log.top.vmax.sex <- lmer(log_top_vmax ~ sex + (1 | ID), data = lab.data)
anova(lmer.log.top.vmax.sex)  
```




> Tukey Test

## don't need to do with sex because it's only 2 groups and because sex doesn't play a big role in overall analysis

```{r}
#Substrate

tukey.lmer.log.side.amax.sub <- emmeans(lmer.log.side.amax.sub, pairwise ~ substrate_type, adjust="tukey", data=lab.data)
tukey.lmer.log.side.amax.sub
plot(tukey.lmer.log.side.amax.sub)
#p-value is significant for D-S but not for D-SD and S-SD

tukey.lmer.log.top.amax.sub <- emmeans(lmer.log.top.amax.sub, pairwise ~ substrate_type, adjust="tukey", data=lab.data)
tukey.lmer.log.top.amax.sub
plot(tukey.lmer.log.top.amax.sub)
#p-value significant for D-S and D-S but not for S-SD
```

> Bipedalism and Velocity/Acceleration

```{r}
#Bipedalism's effect on side velocity
biped.side.vmax <- lme(log10(bipedal_percent+1) ~ log10(side_vmax+1), random = ~1|ID, data = lab.data_omit)
anova(biped.side.vmax)   #p-value < 0.0001   #significant
hist(residuals(biped.side.vmax))
shapiro.test(residuals(biped.side.vmax))  #normal
summary(biped.side.vmax)
performance::r2(biped.side.vmax)   #low R2s

#velocity does affect bipedalism in side videos 
```

```{r}
#Bipedalism's effect on top velocity
biped.top.vmax <- lme(log10(bipedal_percent+1) ~ log10(top_vmax+1), random = ~1|ID, data = lab.data_omit)
anova(biped.top.vmax)   #p-value = 0.0165   #significant
hist(residuals(biped.top.vmax))
shapiro.test(residuals(biped.top.vmax))   #normal
summary(biped.top.vmax)
performance::r2(biped.top.vmax)   #low R2s

#velocity does affect bipedalism in top videos
```

```{r}
#Bipedalism's effect on side acceleration
biped.side.amax <- lme(log10(bipedal_percent+1) ~ log10(side_amax+1), random = ~1|ID, data = lab.data_omit)
anova(biped.side.amax)   #p-value = 0.9311   #not significant
hist(residuals(biped.side.amax))
shapiro.test(residuals(biped.side.amax))  #normal
summary(biped.side.amax)
performance::r2(biped.side.amax)   #low R2s

#acceleration does not affect bipedalism in side videos 
```

```{r}
#Bipedalism's effect on top acceleration
biped.top.amax <- lme(log10(bipedal_percent+1) ~ log10(top_amax+1), random = ~1|ID, data = lab.data_omit)
anova(biped.top.amax)   #p-value = 0.0081   #significant
hist(residuals(biped.top.amax))
shapiro.test(residuals(biped.top.amax))   #normal
summary(biped.top.amax)
performance::r2(biped.top.amax)   #low R2s

#acceleration does affect bipedalism in top videos, but lizard's aren't starting from a standstill so this may be inaccurate
```


> Pauses

```{r}
side.amax.pause <- lme(log_side_amax ~ pauses_side == "0", random = ~1|ID, data = lab.data)
anova(side.amax.pause)   #p-value = 0.2037  #not signficant
hist(residuals(side.amax.pause))
shapiro.test(residuals(side.amax.pause))   #normal
summary(side.amax.pause)
performance::r2(side.amax.pause)   #low R2s

#pauses do not affect acceleration in side videos
```

```{r}
side.amax.pause <- lme(log_side_amax ~ pauses_side, random = ~1|ID, data = lab.data)
anova(side.amax.pause)   #p-value = 0.2341  #not signficant
hist(residuals(side.amax.pause))
shapiro.test(residuals(side.amax.pause))   #normal
summary(side.amax.pause)
performance::r2(side.amax.pause)   #low R2s
```

```{r}
top.amax.pause <- lme(log_top_amax ~ pauses_side)
```

```{r}
lab.data_pause <- lab.data[!is.na(lab.data$pauses_side), ]
```

```{r}
lab.data_side_pause_0 <- read.csv("~/College of Charleston/Thesis/Videos Digitized/Thesis_videos_digitized_condensed_final_10.23.20_removed_non-standstill_starts+NAs+negatives_side_pause_0.csv")

lab.data_side_pause_1 <- read.csv("~/College of Charleston/Thesis/Videos Digitized/Thesis_videos_digitized_condensed_final_10.15.20_removed_non-standstill_starts+NAs+negatives_side_pause_1+.csv")
```

```{r}
#Pauses effect on side acceleration - 0 pauses
lmer.side.amax.pause0 <- lmer(log10(side_amax+1) ~ pauses_side + (1|ID), data = lab.data_side_pause_0)
summary(lmer.side.amax.pause0)
anova(lmer.side.amax.pause0)   #not working for some reason
tukey.lmer.side.amax.pause0 <- emmeans(lmer.side.amax.pause0, pairwise ~ pauses_side, adjust="tukey", data=lab.data_side_pause_0)
tukey.lmer.side.amax.pause0
```

```{r}
#Pauses effect on side acceleration - 1+ pauses
lmer.side.amax.pause1 <- lmer(log10(side_amax+1) ~ pauses_side + (1|ID), data = lab.data_side_pause_1)


#lme() wouldn't work for some reason
```

```{r}
is.numeric(lab.data_side_pause_0$pauses_side)
```


> Sex

```{r}
side.amax.sex <- lmer(log_side_amax ~ log_bipedal_side + substrate_type + sex + (1|ID), data = lab.data)
anova(side.amax.sex)
summary(side.amax.sex)

side.amax.sex_null <- lmer(log_side_amax ~ log_bipedal_side + substrate_type + (1|ID), data = lab.data)
anova(side.amax.sex_null)
summary(side.amax.sex_null)
```

























