---
title: "Untitled"
author: "Kayla Pehl"
date: "10/6/2020"
output: html_document
---

# This dataset does NOT include trials where the lizard did not start from a standstill

```{r}
#Importing data and required packages

lab.data <- read.csv("~/College of Charleston/Thesis/Videos Digitized/Thesis_videos_digitized_condensed_final_10.15.20_removed_non-standstill_starts+NAs+negatives.csv")

#Creating new data frame without N/A's in bipedal percent column    
##148 rows in lab.data and 88 rows in lab.data_omit, so 60 trials didn't contain bipedalism but 88 did 
lab.data_omit <- lab.data[!is.na(lab.data$bipedal_percent), ]

library(ggplot2)
library(rcompanion)
library(lme4)
library(sjstats)
library(lmerTest)
```

```{r}
#Inserting necessary functions
log_top_vmax <- log(lab.data$top_vmax)
log_side_vmax <- log(lab.data$side_vmax)
```


# Glmer and lmer

## lmer is used to fit linear mixed-effect models, so it assumes that the residual error has a Gaussian distribution. If your dependent variable A is a binary outcome (e.g. a yes/no response), then the error distribution is binomial and not Gaussian. In this case you have to use glmer, which allow to fit a generalized linear mixed-effects model: these models include a link function that allows to predict response variables with non-Gaussian distributions.
## glmer -> use when dependent variable (X) is a binary outcome (e.g. yes/no)
### 1 refers to an intercept; means 'add a random intercept for each ID level'

```{r}
lmer(log_top_vmax ~ bipedal_top + (1 | ID), data = lab.data)   

lmer(top_vmax ~ bipedal_top + (1 | ID), data = lab.data)   #standard deviations and intercept is higher for non-log top_vmax than log_top_vmax
```

#Lmer models
```{r}
#side_amax ~ bipedal_side + substrate_type + sex
lmer(side_amax ~ bipedal_side + substrate_type + sex + (1 | ID), data = lab.data)
parameters::p_value(lmer(side_amax ~ bipedal_side + substrate_type + sex +(1 | ID), data = lab.data))  #Finding p-values
```

```{r}
#top_amax ~ bipedal_top + substrate_type + sex
lmer(top_amax ~ bipedal_top + substrate_type + sex + (1 | ID), data = lab.data)
parameters::p_value(lmer(top_amax ~ bipedal_top + substrate_type + sex +(1 | ID), data = lab.data))  #Finding p-values
```

```{r}
#log_side_vmax ~ bipedal_side + substrate_type + sex
summary(lmer(log_side_vmax ~ bipedal_side + substrate_type + sex + (1 | ID), data = lab.data))
parameters::p_value(lmer(log_side_vmax ~ bipedal_side + substrate_type + sex + (1 | ID), data = lab.data))  #Finding p-values
```

```{r}
#log_top_vmax ~ bipedal_top + substrate_type + sex
lmer(log_top_vmax ~ bipedal_top + substrate_type + sex + (1 | ID), data = lab.data)
parameters::p_value(lmer(log_top_vmax ~ bipedal_top + substrate_type + sex +(1 | ID), data = lab.data))  #Finding p-values
```

```{r}
#log_top_vmax ~ bipedal_percent + substrate_type + sex
lmer(log_top_vmax ~ bipedal_percent + substrate_type + sex + (1 | ID), data = lab.data)
parameters::p_value(log_top_vmax ~ bipedal_percent + substrate_type + sex + (1 | ID), data = lab.data)  #Finding p-values  #not working
```

```{r}
#log_side_vmax ~ bipedal_percent + substrate_type + sex
lmer(log_side_vmax ~ bipedal_percent + substrate_type + sex + (1 | ID), data = lab.data)
parameters::p_value(log_side_vmax ~ bipedal_percent + substrate_type + sex + (1 | ID), data = lab.data)  #Finding p-values   #not working
```










