---
title: "Lab data without no standstill starts with 3 lizards removed"
author: "Kayla Pehl"
date: "10/26/2020"
output: html_document
---

# This dataset does NOT include trials where the lizard did not start from a standstill
# It also does not include the 3 lizards (F, N & P) that did not have at least 1 good run on all three substrates saved

```{r}
#Importing data and required packages

lab.data_reduced <- read.csv("~/College of Charleston/Thesis/Videos Digitized/Thesis_videos_digitized_condensed_final_10.26.20_removed_non-standstill_starts+NAs+negatives+3lizards.csv")
#this data set has lizards F, N & P removed - they were the lizards that did not have at least 1 good run on all three substrates saved. 
  
#Creating new data frame without N/A's in bipedal percent column    
##148 rows in lab.data and 88 rows in lab.data_omit, so 60 trials didn't contain bipedalism but 88 did 
lab.data_reduced_omit <- lab.data_reduced[!is.na(lab.data_reduced$bipedal_percent), ]

library(ggplot2)
library(rcompanion)
library(lme4)
library(sjstats)
library(doBy)
library(car)
library(emmeans)
library(multcompView)
library(nlme)
library(lmtest)
```

```{r}
#Checking to see if all data is in correct formats (i.e. factor vs. numeric)    

is.factor(lab.data_reduced$ID)
is.factor(lab.data_reduced$sex)
is.factor(lab.data_reduced$substrate_type)
is.numeric(lab.data_reduced$pauses_side)
is.numeric(lab.data_reduced$pauses_top)
is.factor(lab.data_reduced$bipedal_quad_side)
is.factor(lab.data_reduced$bipedal_quad_top)
is.numeric(lab.data_reduced$bipedal_side)
is.numeric(lab.data_reduced$bipedal_top)
is.numeric(lab.data_reduced$total_video_frames)
is.numeric(lab.data_reduced$bipedal_percent)
is.factor(lab.data_reduced$no_start_from_standstill)
is.factor(lab.data_reduced$stops_before_end_of_racetrack)
is.numeric(lab.data_reduced$calibration)
is.numeric(lab.data_reduced$side_amax)
is.numeric(lab.data_reduced$side_pmax)
is.numeric(lab.data_reduced$side_posmax)
is.numeric(lab.data_reduced$side_spline_tol)
is.numeric(lab.data_reduced$side_tmax)
is.numeric(lab.data_reduced$side_vmax)
is.numeric(lab.data_reduced$top_amax)
is.numeric(lab.data_reduced$top_pmax)
is.numeric(lab.data_reduced$top_posmax)
is.numeric(lab.data_reduced$top_spline_tol)
is.numeric(lab.data_reduced$top_tmax)
is.numeric(lab.data_reduced$top_vmax)

#All are correct. Therefore nothing in dataset needs to change.
```


# Checking for normality

```{r}
#Checking for normality of data: max accelerations and velocities

#histograms
ggplot(data = lab.data_reduced, aes(side_amax))  + 
  geom_histogram(bins = 11, col = ("grey")) +
  labs(title = "Side Amax")                        #slight skew right

ggplot(data = lab.data_reduced, aes(top_amax))  + 
  geom_histogram(bins = 11, col = ("grey")) +
  labs(title = "Top Amax")                         #looks skewed right

ggplot(data = lab.data_reduced, aes(side_vmax))  + 
  geom_histogram(bins = 13, col = ("grey")) +
  labs(title = "Side Vmax")                        #looks fairly normal

ggplot(data = lab.data_reduced, aes(top_vmax))  + 
  geom_histogram(bins = 11, col = ("grey")) +
  labs(title = "Top Vmax")                         #looks fairly normal

#Shapiro-Wilks normality test (p-value >0.05 assumes normality)
shapiro.test(lab.data_reduced$side_amax)
shapiro.test((lab.data_reduced$top_amax))
shapiro.test((lab.data_reduced$side_vmax))
shapiro.test(lab.data_reduced$top_vmax)

#side_amax: not normal
#top_amax: not normal
#side_vmax: normal
#top_vmax: normal
```


```{r}
#Checking for normality of data: time and position

#histograms
ggplot(data = lab.data_reduced, aes(side_tmax))  + 
  geom_histogram(bins = 11, col = ("grey")) +
  labs(title = "Side Tmax")                           #looks skewed right

ggplot(data = lab.data_reduced, aes(top_tmax))  + 
  geom_histogram(bins = 11, col = ("grey")) +
    labs(title = "Top Tmax")                          #looks skewed right

ggplot(data = lab.data_reduced, aes(side_posmax))  + 
  geom_histogram(bins = 13, col = ("grey")) +
    labs(title = "Side Posmax")                       #looks fairly normal

ggplot(data = lab.data_reduced, aes(top_posmax))  + 
  geom_histogram(bins = 30, col = ("grey")) +
    labs(title = "Top Posmax")

#Shapiro-Wilks normality test (p-value >0.05 assumes normality)
shapiro.test(lab.data_reduced$side_tmax)
shapiro.test((lab.data_reduced$top_tmax))
shapiro.test((lab.data_reduced$side_posmax))
shapiro.test(lab.data_reduced$top_posmax)

#side_tmax: not normal
#top_tmax: not normal
#side_posmax: normal
#top_posmax: not normal
```

```{r}
#Checking for normality of data: bipedal_side and bipedal_top

#histograms
ggplot(data = lab.data_reduced_omit, aes(bipedal_side))  + 
  geom_histogram(bins = 20, col = ("grey"))

ggplot(data = lab.data_reduced_omit, aes(bipedal_top))  + 
  geom_histogram(bins = 20, col = ("grey"))

#Shapiro-Wilks normality test (p-value >0.05 assumes normality)
shapiro.test(lab.data_reduced_omit$bipedal_side)
shapiro.test((lab.data_reduced_omit$bipedal_top))

#bipedal_side: normal
#bipedal_top: not normal
```


> Checking assumptions: normality of residuals
## all residuals should be log transformed

```{r}
#Log transforming all variables 
##included log10 and +1 because it forces 0's to be 0's and not something else

log_side_amax <- log10(lab.data_reduced$side_amax +1)
log_top_amax <- log10(lab.data_reduced$top_amax +1)
log_side_vmax <- log10(lab.data_reduced$side_vmax +1)
log_top_vmax <- log10(lab.data_reduced$top_vmax +1)
log_side_posmax <- log10(lab.data_reduced$side_posmax +1)
log_top_posmax <- log10(lab.data_reduced$top_posmax +1)
log_side_tmax <- log10(lab.data_reduced$side_tmax +1)
log_top_tmax <- log10(lab.data_reduced$top_tmax +1)
log_side_pmax <- log10(lab.data_reduced$side_pmax +1)
log_top_pmax <- log10(lab.data_reduced$top_pmax +1)

log_bipedal_side <- log10(lab.data_reduced$bipedal_side +1)
log_bipedal_top <- log10(lab.data_reduced$bipedal_top +1)
log_bipedal_percent <- log10(lab.data_reduced$bipedal_percent +1)
```


#Writing linear models and Residuals

```{r}
#log.side.amax.biped
log.side.amax.biped <- lm(log_side_amax ~ log_bipedal_side, data = lab.data_reduced)
res.log.side.amax.biped <- resid(log.side.amax.biped)
plot(fitted(log.side.amax.biped), res.log.side.amax.biped)

qqnorm(res.log.side.amax.biped)
qqline(res.log.side.amax.biped,
       col = 'red')
shapiro.test(res.log.side.amax.biped)  #normal
plotNormalHistogram(res.log.side.amax.biped)

#lmer model
lmer.log.side.amax.biped <- lmer(log_side_amax ~ log_bipedal_side + substrate_type + sex + (1 | ID), data = lab.data_reduced)
anova(lmer.log.side.amax.biped)
```


```{r} 
#log.top.amax.biped
log.top.amax.biped <- lm(log_top_amax ~ log_bipedal_top, data = lab.data_reduced)
res.log.top.amax.biped <- resid(log.top.amax.biped)

plot(fitted(log.top.amax.biped), res.log.top.amax.biped)
qqnorm(res.log.top.amax.biped)
qqline(res.log.top.amax.biped,
       col = 'red')
shapiro.test(res.log.top.amax.biped)  #normal
plotNormalHistogram(res.log.top.amax.biped)

#lmer model
lmer.log.top.amax.biped <- lmer(log_top_amax ~ log_bipedal_top + substrate_type + sex + (1 | ID), data = lab.data_reduced)
anova(lmer.log.top.amax.biped)
```

```{r}
#log.side.vmax.biped
log.side.vmax.biped <- lm(log_side_vmax ~ bipedal_side, data = lab.data_reduced)
res.log.side.vmax.biped <- resid(log.side.vmax.biped)

plot(fitted(log.side.vmax.biped), res.log.side.vmax.biped)
qqnorm(res.log.side.vmax.biped)
qqline(res.log.side.vmax.biped,
       col = 'red')
shapiro.test(res.log.side.vmax.biped)  #normal
plotNormalHistogram(res.log.side.vmax.biped)

#lmer model
lmer.log.side.vmax.biped <- lmer(log_side_vmax ~ bipedal_side + substrate_type + sex + (1 | ID), data = lab.data_reduced)
anova(lmer.log.side.vmax.biped)
```

```{r}
#log.top.vmax.biped
log.top.vmax.biped <- lm(log_top_vmax ~ bipedal_top, data = lab.data_reduced)
res.log.top.vmax.biped <- resid(log.top.vmax.biped)

plot(fitted(log.top.vmax.biped), res.log.top.vmax.biped)
qqnorm(res.log.top.vmax.biped)
qqline(res.log.top.vmax.biped,
       col = 'red')
shapiro.test(res.log.top.vmax.biped)  #normal
plotNormalHistogram(res.log.top.vmax.biped)

#lmer model
lmer.log.top.vmax.biped <- lmer(log_top_vmax ~ bipedal_top + substrate_type + sex + (1 | ID), data = lab.data_reduced)
anova(lmer.log.top.vmax.biped)
```


> Glmer and lmer

```{r}
#Velocity
lmer.log.side.vmax.sub <- lmer(log_side_vmax ~ substrate_type + (1 | ID), data = lab.data_reduced)
anova(lmer.log.side.vmax.sub)
summary(lmer.log.side.vmax.sub)

lmer.log.top.vmax.sub <- lmer(log_top_vmax ~ substrate_type + (1 | ID), data = lab.data_reduced)
anova(lmer.log.top.vmax.sub)  
```

```{r}
#Acceleration
lmer.log.side.amax.sub <- lmer(log_side_vmax ~ substrate_type + (1 | ID), data = lab.data_reduced)
anova(lmer.log.side.amax.sub)  

lmer.log.top.amax.sub <- lmer(log_top_vmax ~ substrate_type + (1 | ID), data = lab.data_reduced)
anova(lmer.log.top.amax.sub)  
```  

```{r}
#Bipedalism
lmer.log.side.bipedal.sub <- lmer(log_bipedal_side ~ substrate_type + (1 | ID), data = lab.data_reduced)
anova(lmer.log.side.bipedal.sub)  

lmer.log.top.bipedal.sub <- lmer(log_bipedal_top ~ substrate_type + (1 | ID), data = lab.data_reduced)
anova(lmer.log.top.bipedal.sub)  
```


> Tukey Test

```{r}
#Substrate

tukey.lmer.log.side.amax.sub <- emmeans(lmer.log.side.amax.sub, pairwise ~ substrate_type, adjust="tukey", data=lab.data_reduced)
tukey.lmer.log.side.amax.sub
plot(tukey.lmer.log.side.amax.sub)
#p-value is significant for D-S but not for D-SD and S-SD

tukey.lmer.log.top.amax.sub <- emmeans(lmer.log.top.amax.sub, pairwise ~ substrate_type, adjust="tukey", data=lab.data_reduced)
tukey.lmer.log.top.amax.sub
plot(tukey.lmer.log.top.amax.sub)
#p-value significant for D-S and D-S but not for S-SD
```


>Bipedalism and Velocity/Acceleration

```{r}
#velocity's effect on Bipedalism
biped.side.vmax <- lme(log10(bipedal_percent+1) ~ log10(side_vmax+1), random = ~1|ID, data = lab.data_reduced_omit)
anova(biped.side.vmax)   #p-value < 0.0001   #significant
hist(residuals(biped.side.vmax))
shapiro.test(residuals(biped.side.vmax))  #normal
summary(biped.side.vmax)
performance::r2(biped.side.vmax)   #low R2s

#velocity does affect bipedalism in side videos 
```


```{r}
#Bipedalism's effect on side velocity
lab_data_omit <- lab_data[!is.na(lab_data$bipedal_percent), ]
  #lab_data without trials that were run quadrupedally with no bipedalism present

side.vmax.biped <- lme(log10(side_vmax+1) ~ log10(bipedal_percent+1), random = ~1|ID, data = lab_data_omit)
anova(side.vmax.biped)  #p-value < 0.0001
hist(residuals(side.vmax.biped))
shapiro.test(residuals(side.vmax.biped))  #normal
performance::r2(side.vmax.biped)  #C=0.304  #M=0.273

#bipedalism does effect velocity in side videos
```

```{r}
#Bipedalism's effect on top velocity
top.vmax.biped <- lme(log10(top_vmax+1) ~ log10(bipedal_percent+1), random = ~1|ID, data = lab_data_omit)
anova(top.vmax.biped)  #p-value < 0.0736
hist(residuals(top.vmax.biped))
shapiro.test(residuals(top.vmax.biped))  #normal
performance::r2(top.vmax.biped)  #C=0.425  #M=0.038

#bipedalsim does not effect velocity in top videos
```

```{r}
#Bipedalism's effect on side acceleration
side.amax.biped <- lme(log10(side_amax+1) ~ log10(bipedal_percent+1), random = ~1|ID, data = lab_data_omit)
anova(side.amax.biped)  #p-value < 0.5229
hist(residuals(side.amax.biped))
shapiro.test(residuals(side.amax.biped))  #normal
performance::r2(side.amax.biped)  #C=0.119  #M=0.006

#Bipedalism does not effect acceleration in side videos
```


```{r}
#Velocity's effect on Bipedalism
biped.top.vmax <- lme(log10(bipedal_percent+1) ~ log10(top_vmax+1), random = ~1|ID, data = lab.data_reduced_omit)
anova(biped.top.vmax)   #p-value = 0.0693   #not significant
hist(residuals(biped.top.vmax))
shapiro.test(residuals(biped.top.vmax))   #normal
summary(biped.top.vmax)
performance::r2(biped.top.vmax)   #low R2s

#velocity does affect bipedalism in top videos
```

```{r}
#Acceleration's effect on Bipedalism
biped.side.amax <- lme(log10(bipedal_percent+1) ~ log10(side_amax+1), random = ~1|ID, data = lab.data_reduced_omit)
anova(biped.side.amax)   #p-value = 0.7871   #not significant
hist(residuals(biped.side.amax))
shapiro.test(residuals(biped.side.amax))  #normal
summary(biped.side.amax)
performance::r2(biped.side.amax)   #low R2s

#acceleration does not affect bipedalism in side videos 
```

```{r}
#Bipedalism's effect on top acceleration
biped.top.amax <- lme(log10(bipedal_percent+1) ~ log10(top_amax+1), random = ~1|ID, data = lab.data_reduced_omit)
anova(biped.top.amax)   #p-value = 0.022   #significant
hist(residuals(biped.top.amax))
shapiro.test(residuals(biped.top.amax))   #normal
summary(biped.top.amax)
performance::r2(biped.top.amax)   #low R2s

#acceleration does affect bipedalism in top videos, but lizard's aren't starting from a standstill so this may be inaccurate
```

