---
title: "Untitled"
author: "Kayla Pehl"
date: "10/6/2020"
output: html_document
---

# This dataset does NOT include trials where the lizard did not start from a standstill

```{r}
#Importing data and required packages

lab.data <- read.csv("~/College of Charleston/Thesis/Videos Digitized/Thesis_videos_digitized_condensed_final_10.08.20_removed_non-standstill_starts+NA.csv")
library(ggplot2)
library(rcompanion)
library(lme4)
library(sjstats)
library(lmerTest)
```

```{r}
#Inserting necessary functions
log_top_vmax <- log(lab.data$top_vmax)
log_side_vmax <- log(lab.data$side_vmax)
```



# Glmer and lmer

## lmer is used to fit linear mixed-effect models, so it assumes that the residual error has a Gaussian distribution. If your dependent variable A is a binary outcome (e.g. a yes/no response), then the error distribution is binomial and not Gaussian. In this case you have to use glmer, which allow to fit a generalized linear mixed-effects model: these models include a link function that allows to predict response variables with non-Gaussian distributions.
## glmer -> use when dependent variable (X) is a binary outcome (e.g. yes/no)
### 1 refers to an intercept; means 'add a random intercept for each ID level'

```{r}
lmer(log_top_vmax ~ bipedal_top + (1 | ID), data = lab.data)   
```

```{r}
#Mixed model lmer 
summary(lmer(log_side_vmax ~ bipedal_side + (1 | ID), data = lab.data))

#Finding p-value in lmer 
sjstats::p_value(lmer(log_side_vmax ~ bipedal_side + (1 | ID), data = lab.data))
```


```{r}
#Mixed model lmer
summary(lmer(log_top_vmax ~ bipedal_top + (1 | ID), data = lab.data))

#Finding p-value in lmer 
sjstats::p_value(lmer(log_top_vmax ~ bipedal_top + (1 | ID), data = lab.data))
```


```{r}
summary(lmer(log_side_vmax ~ substrate_type + (1 | ID), data = lab.data))
```


```{r}
summary(lmer(log_side_vmax ~ substrate_type*sex + (1 | ID), data = lab.data))
```

```{r}
summary(lmer(log_top_vmax ~ substrate_type*sex + (1 | ID), data = lab.data))
#Sand is significant but not SD

summary(lmer(log_top_vmax ~ substrate_type + (1 | ID), data = lab.data))
#Sand and SD are both significant, even though SD wasn't significant when sex was included

summary(lmer(log_top_vmax ~ sex + (1 | ID), data = lab.data))
#sex wasn't significant in any of the lmers
```


```{r}

```













































