---
title: "Lab_data_w/o_no-standstill_starts"
author: "Kayla Pehl"
date: "10/1/2020"
output: html_document
---

# This dataset does NOT include trials where the lizard did not start from a standstill

```{r}
#Importing data and required packages

data <- read.csv("~/College of Charleston/Thesis/Videos Digitized/Thesis_videos_digitized_condensed_final_10.01.20_removed_non-standstill_starts.csv")
library(ggplot2)
library(rcompanion)
library(lme4)
```

```{r}
#Side max acceleration max per substrate 

ggplot(data = data, aes(side_amax))  + 
  geom_histogram(fill="orange", bins = 20, binwidth = 10) +
  facet_wrap(~substrate_type) +
  labs(title = "Side max acceleration per substrate")


#Side max acceleration max per sex

ggplot(data = data, aes(side_amax))  + 
  geom_histogram(fill="orange", bins = 20, binwidth = 5) +
  facet_wrap(~sex) +
  labs(title = "Side max acceleration per sex")
```

```{r}
#Top max acceleration max per substrate 

ggplot(data = data, aes(top_amax))  + 
  geom_histogram(fill="purple", bins = 30, binwidth = 10) +
  facet_wrap(~substrate_type) +
  labs(title = "Top max acceleration per substrate")


#Top max acceleration max per sex

ggplot(data = data, aes(top_amax))  + 
  geom_histogram(fill="purple", bins = 30, binwidth = 5) +
  facet_wrap(~sex) +
  labs(title = "Top max acceleration per sex")
```


```{r}
#Bipedalism percentage per sex Histogram
ggplot(data = data, aes(as.numeric(bipedal_percent)))  + 
  geom_histogram(fill="green", bins = 20, binwidth = 10) +
  facet_wrap(~sex)

#Bipedalism percentage per substrate Histogram
ggplot(data = data, aes(as.numeric(bipedal_percent)))  + 
  geom_histogram(fill="green", bins = 20, binwidth = 10) +
  facet_wrap(~substrate_type)
```


```{r}
#Boxplots of Bipedalism on acceleration 

ggplot(data = data, aes(x=bipedal_quad_side, y=side_amax)) +
  geom_boxplot() +
    labs(title = "Bipedalism effect on max acceleration (side)")

ggplot(data = data, aes(x=bipedal_quad_top, y=top_amax)) +
  geom_boxplot() +
    labs(title = "Bipedalism effect on max acceleration (top)")

#Boxplots of Bipedalism on velocity

ggplot(data = data, aes(x=bipedal_quad_side, y=side_vmax)) +
  geom_boxplot() +
    labs(title = "Bipedalism effect on max velocity (side)")

ggplot(data = data, aes(x=bipedal_quad_top, y=top_vmax)) +
  geom_boxplot() +
    labs(title = "Bipedalism effect on max velocity (top)")
```

```{r}
table(data$bipedal_quad_side)
table(data$bipedal_quad_top)
```


```{r}
#Summary of data
summary(data)
```

```{r}
summary(lm(side_amax~sex, data = data))
```

```{r}
#Checking for normality of data: max accelerations and velocities

#historgrams
ggplot(data = data, aes(side_amax))  + 
  geom_histogram(bins = 11, col = ("grey"))

ggplot(data = data, aes(top_amax))  + 
  geom_histogram(bins = 11, col = ("grey"))

ggplot(data = data, aes(side_vmax))  + 
  geom_histogram(bins = 13, col = ("grey"))

ggplot(data = data, aes(top_vmax))  + 
  geom_histogram(bins = 11, col = ("grey"))

#Shapiro-Wilks normality test (p-value <0.05 is normal)
shapiro.test(data$side_amax)
shapiro.test((data$top_amax))
shapiro.test((data$side_vmax))
shapiro.test(data$top_vmax)

#side_amax: normal
#top_amax: normal
#side_vmax: not normal
#top_vmax: not normal
```


```{r}
#Checking for normality of data: time and position

#historgrams
ggplot(data = data, aes(side_tmax))  + 
  geom_histogram(bins = 11, col = ("grey"))

ggplot(data = data, aes(top_tmax))  + 
  geom_histogram(bins = 11, col = ("grey"))

ggplot(data = data, aes(side_posmax))  + 
  geom_histogram(bins = 13, col = ("grey"))

ggplot(data = data, aes(top_posmax))  + 
  geom_histogram(bins = 11, col = ("grey"))

#Shapiro-Wilks normality test (p-value <0.05 is normal)
shapiro.test(data$side_tmax)
shapiro.test((data$top_tmax))
shapiro.test((data$side_posmax))
shapiro.test(data$top_posmax)

#side_tmax: normal
#top_tmax: normal
#side_posmax: not normal
#top_posmax: normal
```


```{r}
#Top camera max acceleration vs. bipedalism in side

top.amax.biped <- lm(top_amax ~ bipedal_top, data = data)
res.top.amax.biped <- resid(top.amax.biped)
plot(fitted(top.amax.biped), res.top.amax.biped)
abline(0,0)

qqnorm(res.top.amax.biped)
qqline(res.top.amax.biped,
       col = 'red')
plot(density(res.top.amax.biped))

plot(top.amax.biped)
```

```{r}
#Side camera max acceleration vs. bipedalism in side
side.amax.biped <- lm(side_amax ~ bipedal_side, data = data)
res.side.amax.biped <- resid(side.amax.biped)
plot(fitted(side.amax.biped), res.side.amax.biped)

qqnorm(res.side.amax.biped)
qqline(res.side.amax.biped,
       col = 'red')
plot(density(res.side.amax.biped))

plot(side.amax.biped)
```

```{r}
#Side camera max velocity vs. bipedalism in side
side.vmax.biped <- lm(side_vmax ~ bipedal_side, data = data)
res.side.vmax.biped <- resid(side.vmax.biped)
plot(fitted(side.vmax.biped), res.side.vmax.biped)

qqnorm(res.side.vmax.biped)
qqline(res.side.vmax.biped,
       col = 'red')
plot(density(res.side.vmax.biped))

plot(side.vmax.biped)
```

```{r}
#Log transform side camera max velocity vs. bipedalism in side

log_side_vmax <- log(data$side_vmax)
log.side.vmax.biped <- lm(log_side_vmax ~ bipedal_side, data = data)
res.log.side.vmax.biped <- resid(log.side.vmax.biped)
plot(fitted(log.side.vmax.biped), res.log.side.vmax.biped)

qqnorm(res.log.side.vmax.biped)
qqline(res.log.side.vmax.biped,
       col = 'red')
plot(density(res.log.side.vmax.biped))

plot(log.side.vmax.biped)

ggplot(data = data, aes(log_side_vmax))  + 
  geom_histogram(bins = 13, col = ("grey"))    #skewed left

#Q-Q plots of regular and log-transformed side vmax data look basically the same to me. I'm not sure if log-transforming did a lot for the data.
```

```{r}
#Comparing side vmax data from original with log-transformed

#histograms
plotNormalHistogram(data$side_vmax)  #appears pretty symmetrical 
plotNormalHistogram(log_side_vmax)  #skew left

#Shapiro-Wilks test for normality
shapiro.test(data$side_vmax)  #not normal
shapiro.test(log_side_vmax)  #normal
```


```{r}
#Log transform top camera max velocity vs. bipedalism in top

log_top_vmax <- log(data$top_vmax)
log.top.vmax.biped <- lm(log_top_vmax ~ bipedal_top, data = data)
res.log.top.vmax.biped <- resid(log.top.vmax.biped)
plot(fitted(log.top.vmax.biped), res.log.top.vmax.biped)

qqnorm(res.log.top.vmax.biped)
qqline(res.log.top.vmax.biped,
       col = 'red')
plot(density(res.log.top.vmax.biped))

plot(log.top.vmax.biped)
```

```{r}
#Comparing top vmax data from original with log-transformed

#histograms
plotNormalHistogram(data$top_vmax)  #appears pretty symmetrical 
plotNormalHistogram(log_top_vmax)  #skew left

#Shapiro-Wilks test for normality
shapiro.test(data$top_vmax)  #not normal
shapiro.test(log_top_vmax)  #normal
```

```{r}
lmer(log_top_vmax ~ bipedal_top + (1 | ID), data = data)   #1 refers to an intercept; means 'add a random intercept for each ID level'
```

## lmer is used to fit linear mixed-effect models, so it assumes that the residual error has a Gaussian distribution. If your dependent variable A is a binary outcome (e.g. a yes/no response), then the error distribution is binomial and not Gaussian. In this case you have to use glmer, which allow to fit a generalized linear mixed-effects model: these models include a link function that allows to predict response variables with non-Gaussian distributions.
## glmer -> use when dependent variable (X) is a binary outcome (e.g. yes/no)


