---
title: "Lizard_Running_Power_Analysis"
author: "Kayla Pehl"
date: "3/9/2020"
output: html_document
---

# Before running the code below, please read the "README.md" located in the repository.


POWER ANALYSIS

```{r}
# Install "pwr" package for power analysis prior to running code. 
# Loading required packages
library(pwr)   
```


Power analysis t-test(n = sample size, d = effect size, sig.level = significance 
level, power = power, type = t-test type)

# d = (mean group 1 - mean group 2) / standard deviation

```{r}
# Ballinger et al. (1979) running speed of Racerunners on a racetrack:
running_time <- c(1.63, 1.55, 1.39, 1.26, 1.53, 1.40, 2.28, 1.28, 1.95, 1.62)
# Power Analysis Constants:
# set percent of effect
eff_pct <- 0.15
cnt_mean <- mean(running_time)  #control mean
trt_mean <- cnt_mean + (cnt_mean * eff_pct)   #treatment mean
cnt_sd <- sd(running_time)  #standard deviation = variance
d <- ((cnt_mean - trt_mean) / cnt_sd)  #d for power analysis
cnt_mean
trt_mean
cnt_sd
d
```


> Running Power Analyses

```{r}
pwr.t.test(n = NULL, d = d, sig.level = 0.05, power = 0.80, type = "one.sample", 
           alternative = "two.sided") 
```

```{r}
pwr.t.test(n = NULL, d = d, sig.level = 0.01, power = 0.80, type = "one.sample", 
           alternative = "two.sided")
```

```{r}
pwr.t.test(n = NULL, d = d, sig.level = 0.05, power = 0.90, type = "one.sample", 
           alternative = "two.sided")
```

```{r}
pwr.t.test(n = NULL, d = d, sig.level = 0.01, power = 0.90, type = "one.sample",
           alternative = "two.sided")
```

> Tried a couple of different effect sizes (d). When using a significance level 
of 0.05, a power level of 0.80, and an effect of 0.5 (50%), I got a sample size 
of 3, which seemed very small. Using the same parameters except for effect size, 
which I changed to 0.1 (10%), I got a sample size of 32. Changing the effect 
size to 0.15 (15%) produced a sample size of 15, and using an effect size of 0.2 
gave a sample size of 10. These different effect sizes show me that I would like 
a sample size of at least 10, and up to possibly 30, depending on how variable 
the lizards are running on the substrates.  
> I changed the significance level and power level in the power analyses above
to get a broader sense of what type of sample size I would need depending on 
how rigorous I wanted my analysis to be. The lower the significance level and 
the higher the power, the greater sample size required.