---
title: 'Question 2: How does "available" habitat affect Racerunner locomotion?'
author: "Kayla Pehl"
date: "10/16/2020"
output: html_document
---

Question 2: How does "available" habitat affect Racerunner locomotion?
 Ho: Racerunners run the same on all substrates.
 Ha: Racerunners run fastest on sand compared to debris and sand/debris.

> Data used to answer this question:
   > lab experiment data
      > velocity vs. substrate type - lme
      > acceleration vs. substrate type - lme
      > bipedalism vs. substrate type - lme
         > does bipedalism affect velocity? bipedalism vs velocity - lme
      > power vs. substrate type ? - lme
      > time vs. substrate type ? - lme
   > McElroy habitat data --> Q1
      > speed vs. ground clutter crossed -- are speeds faster when more/less ground clutter is crossed?
          > will need to distinguish which category (S/SD/D) amount of ground clutter (0-20) corresponds with ??
          > lmer(speed ~ ground_clutter_crossed + (1 | ID))      

Results
   Racerunners ran fastest on debris and slowest on sand. Sand/debris is treated more like sand than debris.
   Mixed was the most common substrate in the field.
   Acceleration doesn't appear to be affected by substrate type


RE-DO USING NEW DATA SET!!!

```{r}
#Importing data and required packages

lab.data <- read.csv("~/College of Charleston/Thesis/Videos Digitized/Thesis_videos_digitized_condensed_final_10.15.20_removed_non-standstill_starts+NAs+negatives.csv")

#Creating new data frame without N/A's in bipedal percent column    
##148 rows in lab.data and 88 rows in lab.data_omit, so 60 trials didn't contain bipedalism but 88 did 
lab.data_omit <- lab.data[!is.na(lab.data$bipedal_percent), ]

library(ggplot2)
library(rcompanion)
library(lme4)
library(sjstats)
library(doBy)
library(car)
library(emmeans)
library(multcompView)
library(nlme)
library(lmtest)
```


> Lab Experiment

# Velocity vs. substrate type

```{r}
#Velocity vs Substrate Type - Side
lme.log.side.vmax.sub <- lme(log10(side_vmax+1) ~ substrate_type, random = ~1|ID, data = lab.data)
anova(lme.log.side.vmax.sub)  #p-value=0.0037  #significant
hist(residuals(lme.log.side.vmax.sub))
shapiro.test(residuals(lme.log.side.vmax.sub))  #normal
summary(lme.log.side.vmax.sub)
performance::r2(lme.log.side.vmax.sub)  #low R2s
lme.log.side.vmax.sub.tukey <- emmeans(lme.log.side.vmax.sub, pairwise ~ substrate_type, adjust="tukey", data=lab.data)

sqrt(.396)
lme.log.side.vmax.sub.tukey  #significant differences between D-S but none of the others
```

```{r}
#Velocity vs Substrate Type - Top
lme.log.top.vmax.sub <- lme(log10(top_vmax+1) ~ substrate_type, random = ~1|ID, data = lab.data)
anova(lme.log.top.vmax.sub)  #p-value=0.0005  #significant
hist(residuals(lme.log.top.vmax.sub))
shapiro.test(residuals(lme.log.top.vmax.sub))  # not normal
summary(lme.log.top.vmax.sub)
performance::r2(lme.log.top.vmax.sub)  #low R2s
lme.log.top.vmax.sub.tukey <- emmeans(lme.log.top.vmax.sub, pairwise ~ substrate_type, adjust="tukey", data=lab.data)
lme.log.top.vmax.sub.tukey  #significant differences between D-S and D-SD but not S-SD

#DATA IS NOT NORMAL -- TRANSFORM
```

```{r}
#Velocity vs Substrate Type - Top (transformed)
lme.top.vmax.sub <- lme(top_vmax+1 ~ substrate_type, random = ~1|ID, data = lab.data)
anova(lme.top.vmax.sub)  #p-value=  #
hist(residuals(lme.top.vmax.sub))
shapiro.test(residuals(lme.top.vmax.sub))  # 
summary(lme.top.vmax.sub)
performance::r2(lme.top.vmax.sub)  #
lme.top.vmax.sub.tukey <- emmeans(lme.top.vmax.sub, pairwise ~ substrate_type, adjust="tukey", data=lab.data)
lme.top.vmax.sub.tukey  #
```


# Acceleration vs. substrate type

```{r}
#Acceleration vs Substrate Type - Side
lme.log.side.amax.sub <- lme(log10(side_amax+1) ~ substrate_type, random = ~1|ID, data = lab.data)
anova(lme.log.side.amax.sub)  #p-value=0.8524  #not significant
hist(residuals(lme.log.side.amax.sub))
shapiro.test(residuals(lme.log.side.amax.sub))  #normal
summary(lme.log.side.amax.sub)
performance::r2(lme.log.side.amax.sub)
lme.log.side.amax.sub.tukey <- emmeans(lme.log.side.amax.sub, pairwise ~ substrate_type, adjust="tukey", data=lab.data)
lme.log.side.amax.sub.tukey  #no significant differences in any substrate's acceleration
```


#Bipedalism vs. substrate type

```{r}
#Bipedalism vs Substrate Type
lme.biped.sub <- lme(sqrt(bipedal_percent) ~ substrate_type, random = ~1|ID, data = lab.data_omit)
anova(lme.biped.sub)  #p-value=0.0246  #significant
hist(residuals(lme.biped.sub))  
shapiro.test(residuals(lme.biped.sub))   #normal
summary(lme.biped.sub)
performance::r2(lme.biped.sub)  #conditional: 0.427, marginal: 0.06  #good!
lme.biped.sub.tukey <- emmeans(lme.biped.sub, pairwise ~ substrate_type, adjust="tukey", data=lab.data_omit)
lme.biped.sub.tukey  #significant differences in D-S but not D-SD or S-SD
```



