---
title: 'Question 1: Do Racerunners use their habitat randomly or non-randomly?'
author: "Kayla Pehl"
date: "10/16/2020"
output: html_document
---

Question 1: Do Racerunners use their habitat randomly or non-randomly?
 Ho: Racerunners use their habitat randomly.
 Ha: Racerunners use their habitat non-randomly.

> Data used for this question: 
    > habitat transect data
        > percent occurrance of each type of substrate (habitat data) -- what is most common?
        > does cover effect it?
    > McElroy habitat data
        > speed vs. ground clutter crossed -- are speeds faster when more/less ground clutter is crossed?
           > lmer(speed ~ ground_clutter_crossed + (1 | ID))
        > ground clutter crossed effect on bipedalism -- probably can't do because there aren't any definite 'Y's'
    > lab data  
        > how often bipedalism occurs on different substrates ?
          > check histograms
        
Results:
    There is a slight upward trend in McElroy habitat data between speed and ground clutter crossed, which suggests that Racerunners run slightly faster
    when there are more obstacles in their path.
  Transect data: mixed (specifically heavy mixed) was most common substrate in field.
  Racerunners are faster and run bipedally more on debris than sand or sand/debris.
  Racerunners use their habitat non-randomly.


RE-DO USING NEW DATA SET!!!

```{r}
#Loading necessary packages
library(ggplot2)
library(rcompanion)
library(lme4)
library(sjstats)
library(doBy)
library(car)
library(emmeans)
library(multcompView)
library(nlme)
library(lmtest)
```

```{r}
#Importing datasets

McElroy_Habitat_Data2 <- read.csv("~/College of Charleston/Thesis/Data/McElroy Habitat Data/McElroy_Habitat_Data2.csv")
mcelroy_habitat_data_omit <-  read.csv("~/College of Charleston/Thesis/Data/McElroy Habitat Data/McElroy_Habitat_Data_omit.csv")  #data set minus runs where bipedalism was unknown
McElroy_Habitat_Data2_heavy <- read.csv("~/College of Charleston/Thesis/Data/McElroy Habitat Data/McElroy_Habitat_Data2_heavy.csv")  #10-20 obstacles
McElroy_Habitat_Data2_light <- read.csv("~/College of Charleston/Thesis/Data/McElroy Habitat Data/McElroy_Habitat_Data2_light.csv")  #0-9 obstacles
field_transects <- read.csv("~/College of Charleston/Thesis/Data/Field_Transects_Substrates.csv")  #field transect data
field_transects_totals <-  read.csv("~/College of Charleston/Thesis/Data/Field_Transects_Substrate_Totals.csv")  #field transect substrate totals
field_transects_totals_condensed <- read.csv("~/College of Charleston/Thesis/Data/Field_Transects_Substrate_Totals_condensed.csv")
lab.data <- read.csv("~/College of Charleston/Thesis/Videos Digitized/Thesis_videos_digitized_condensed_final_10.15.20_removed_non-standstill_starts+NAs+negatives.csv")
field_obstacles <- read.csv("~/College of Charleston/Thesis/Data/field_obstacles.csv")
field_transects_substrate_heaviness <- read.csv("~/College of Charleston/Thesis/Data/Field Transects/Field_Transects_Substrate_heaviness.csv")
```


```{r}
field_transects_totals
#heavy mixed was most common substrate type in transect study
field_transects_totals_condensed
#mixed was most common grouped susbtrate type
field_transects_substrate_heaviness
#lizards therefore have to be at least somewhat good at running on heavy mixed because there's so much of it in their habitat
```

```{r}
#Bipedalism percentage per substrate and sex Histogram
ggplot(data = lab.data_omit, aes(as.numeric(bipedal_percent)))  + 
  geom_histogram(fill="dark green", bins = 20, binwidth = 10) +
  facet_wrap(~substrate_type + sex) +
  facet_grid(sex ~ substrate_type)

#mean, median & range of bipedal percentage separated into substrate type
summaryBy(bipedal_percent ~ substrate_type + sex, data = lab.data_omit,
          FUN = list(mean, median, min, max))

#males are bipedal more often on every substrate than females
``` 

```{r}
speed_bipedal <- lmer(speed ~ bipedal + (1 | ï..ID), data = mcelroy_habitat_data_omit)
anova(speed_bipedal)
summary(speed_bipedal)
```


> Work from meeting with Eric

```{r}
#Ground Clutter Crossed vs. Speed
speed_ground_cover <- lmer(speed ~ ground_clutter_crossed + (1 | ï..ID), data = McElroy_Habitat_Data2)
anova(speed_ground_cover)
summary(speed_ground_cover)

speed_ground_cover_null <- lmer(speed ~ (1 | ï..ID), data = McElroy_Habitat_Data2)

plot(McElroy_Habitat_Data2$ground_clutter_crossed, McElroy_Habitat_Data2$speed)

ggplot(McElroy_Habitat_Data2, aes(ground_clutter_crossed, speed)) +
  geom_point() + 
  geom_smooth(method = 'lm') +
  labs(title = "Ground Clutter Crossed vs Speed in Habitat Data")

#There is a slight upward trend in speed as number of ground clutter crossed increases. This suggests that Racerunners run slightly faster when there are more obstacles in their path. 
```
  

```{r}
#ensuring ground_clutter_crossed is still numeric (it wasn't before, but we changed it, hence the "2" at the end of the dataset name)
is.numeric(McElroy_Habitat_Data2$ground_clutter_crossed)
```

```{r}
#Likelihood ratio test to compare the fit of 2 models
speed_ground_cover_null <- lmer(speed ~ (1 | ï..ID), data = McElroy_Habitat_Data2)
lrtest(speed_ground_cover, speed_ground_cover_null)  #p-value: 0.6112
lrtest(speed_ground_cover_null, speed_ground_cover)   #p-value: 0.6112

#creating a null function allows you to see if adding in a variable does a lot to change the variation in the analysis
#changing the order of the lrtest doesn't change the outcome of the likelihood test
#Ho: the smaller model is the "best" model. It's rejected when the p-value is large (like it is here), which means that the larger model is a significant improvement over the smaller one.   https://www.statisticshowto.com/likelihood-ratio-tests/
  #This means that speed ground clutter crossed does impact speed
```

```{r}
performance::r2(speed_ground_cover)
performance::r2(speed_ground_cover_null)

#conditional and marginal R2s are both higher for regular model (non-null)
```



```{r}
anova(speed_ground_cover)
summary(speed_ground_cover)

lm(speed ~ ground_clutter_crossed, data = McElroy_Habitat_Data2)
anova(lm(speed ~ ground_clutter_crossed, data = McElroy_Habitat_Data2))
anova(lm(log10(speed) ~ log10(ground_clutter_crossed +1) , data = McElroy_Habitat_Data2))
```

```{r}
#Ground Clutter Crossed vs. Speed Lmer Models

#Regular Model
speed.clutter <- lme(speed ~ ground_clutter_crossed, random= ~1|ï..ID, data = McElroy_Habitat_Data2, na.action=na.omit)
anova(speed.clutter)   #p-value = 0.0308  #significant
hist(residuals(speed.clutter))
shapiro.test(residuals(speed.clutter))  #not normal
summary(speed.clutter)
performance::r2(speed.clutter)   #low R2s

#Log-transformed Model
log.speed.clutter <- lme(log10(speed) ~ log10(ground_clutter_crossed+1), random= ~1|ï..ID, data = McElroy_Habitat_Data2, na.action=na.omit)
anova(log.speed.clutter)   #p-value = 0.0412  #significant
hist(residuals(log.speed.clutter))
shapiro.test(residuals(log.speed.clutter))  #normal p=0.2585
performance::r2(log.speed.clutter)  #low R2s

plot(log10(McElroy_Habitat_Data2$ground_clutter_crossed +1), log10(McElroy_Habitat_Data2$speed))
ggplot(McElroy_Habitat_Data2, aes(log10(ground_clutter_crossed+1), log10(speed)) )+
  geom_point() + 
  geom_smooth(method = 'lm') +
  labs(title = "Log Ground Clutter Crossed vs Log Speed in Habitat Data")

#have to use log-transformed model because non-transformed is not normal
```

# R2 determines how much a variable's behavior can explain the behavior of another variable. A measure 0f 70% or more means that the behavior of the dependent variable is highly explained by the behavior of the independent variable being studied.
  # marginal R2 describes the proportion of variance explained by the fixed factor(s) alone.
  # conditoinal R2 describes the proportion of variance explained by both the fixed and random factors.



> post-meeting

```{r}
lsmeans(random, pairwise ~ ground_clutter_crossed, adjust = "tukey", data=McElroy_Habitat_Data2)
```

```{r}
#Cover vs distance plot of transect data
ggplot(data = field_transects_totals, aes(x=cover_space_cm, y=distance_cm, color = substrate)) +
  geom_point() +
  labs(title = "Transect Data: Cover (cm) vs. Distance (cm)")
```

```{r}
#Distance of Movement vs. Speed
speed_distance_move <- lmer(speed ~ distance_movement + (1 | ï..ID), data = McElroy_Habitat_Data2)
anova(speed_distance_move)
summary(speed_distance_move)

speed_distance_move_null <- lmer(speed ~ (1 | ï..ID), data = McElroy_Habitat_Data2)

ggplot(McElroy_Habitat_Data2, aes(distance_movement, speed)) +
  geom_point() + 
  geom_smooth(method = 'lm') +
  labs(title = "Distance of Movement vs Speed in Habitat Data")

#There is a slight upward trend in speed as the distance of each movement increases. This suggests that Racerunners run slightly faster when they run longer distances. 
```


> Bipedalism on difference substrates

```{r}
#Bipedalism percentage per substrate and sex Histogram
ggplot(data = lab.data_omit, aes(as.numeric(bipedal_percent)))  + 
  geom_histogram(fill="dark green", bins = 20, binwidth = 10) +
  facet_wrap(~substrate_type + sex) +
  facet_grid(sex ~ substrate_type) +
  labs(title = "Histogram of Bipedalism Percentage per Substrate")

#mean, median & range of bipedal percentage separated into substrate type
summaryBy(bipedal_percent ~ substrate_type + sex, data = lab.data_omit,
          FUN = list(mean, median, min, max))

#males are bipedal more often on every substrate than females

##for final manuscript, change plot titles to actual names instead of letters
### https://www.zevross.com/blog/2019/04/02/easy-multi-panel-plots-in-r-using-facet_wrap-and-facet_grid-from-ggplot2/ 
```

```{r}
#Histogram of bipedal percent without NA's separated into substrate type and sex
ggplot(data = lab.data_omit, aes(as.numeric(bipedal_percent), col=substrate_type, fill=substrate_type, alpha=sex)) +
  geom_histogram(bins = 20, binwidth = 8) +
  facet_wrap(~substrate_type) +
  labs(title = "Histogram of Bipedal Percent (without N/As) Separated by Substrate Type and Sex")

#Boxplot of bipedal percent without N/A's separated by substrate type and sex
ggplot(data = lab.data_omit, aes(substrate_type, as.numeric(bipedal_percent), fill = sex)) +
  geom_boxplot() +
  labs(title = "Boxplot of Bipedal Percent (without N/As) by Substrate Type and Sex")

#Histogram of bipedal percent without N/A's broken into substrate type
ggplot(data = lab.data_omit, aes(as.numeric(bipedal_percent), col=substrate_type, fill=substrate_type)) +
  geom_histogram(col = "white") + 
  labs(title = "Histogram of Bipedal Percent (without N/As) Broken Down by Substrate Type")
```


# Trying to compare heavy vs. light obstacles crossed in McElroy habitat data. However, there probably is not enough heavy data to do a good analysis on. Therefore comparing this to the light data is probably pointless.
# 0-9 obstsacles = light
# 10-20 obstacles = heavy
```{r}
#Ground clutter crossed (heavy) vs. speed
speed.clutter.heavy <- lme(log10(speed+1) ~ log10(ground_clutter_crossed+1), random= ~1|ID, data = McElroy_Habitat_Data2_heavy, na.action=na.omit)
anova(speed.clutter.heavy)  #p-value = 02644  #not significant
hist(residuals(speed.clutter))
shapiro.test(residuals(speed.clutter.heavy))  #normal
summary(speed.clutter.heavy)
performance::r2(speed.clutter.heavy)   #low R2s

#Ground clutter crossed (light) vs. speed
speed.clutter.light <- lme(log10(speed+1) ~ log(ground_clutter_crossed+1), random= ~1|ID, data = McElroy_Habitat_Data2_light, na.action=na.omit)
anova(speed.clutter.light)  #p-value = 0.0309  # significant
hist(residuals(speed.clutter))
shapiro.test(residuals(speed.clutter.light))  #normal
summary(speed.clutter.light)
performance::r2(speed.clutter.light)   #low R2s
```


> Velocity per Substrate

```{r}
#Top max velocity per substrate 
ggplot(data = lab.data, aes(top_vmax))  + 
  geom_histogram(fill="purple", bins = 10) +
  facet_wrap(~substrate_type) +
  labs(title = "Top max velocity per substrate")

#mean, median & range of top max velocity separated into substrate types
summaryBy(top_vmax ~ substrate_type, data = lab.data,
          FUN = list(mean, median, min, max))
```

```{r}
#Top max velocity per sex
ggplot(data = lab.data, aes(top_vmax))  + 
  geom_histogram(fill="purple", bins = 10) +
  facet_wrap(~sex) +
  labs(title = "Top max velocity per sex")

#mean, median & range of top max velocity separated into sex
summaryBy(top_vmax ~ sex, data = lab.data,
          FUN = list(mean, median, min, max))
```

```{r}
#Velocity

log_side_vmax <- log10(lab.data$side_vmax +1)
log_top_vmax <- log10(lab.data$top_vmax +1)

lmer.log.side.vmax.sub <- lmer(log_side_vmax ~ substrate_type + (1 | ID), data = lab.data)
anova(lmer.log.side.vmax.sub)
summary(lmer.log.side.vmax.sub)

lmer.log.top.vmax.sub <- lmer(log_top_vmax ~ substrate_type + (1 | ID), data = lab.data)
anova(lmer.log.top.vmax.sub)  


``` 
  