---
title: 'Question 1: Do Racerunners use their habitat randomly or non-randomly?'
author: "Kayla Pehl"
date: "10/16/2020"
output: html_document
---

Question 1: Do Racerunners use their habitat randomly or non-randomly?
 Ho: Racerunners use their habitat randomly.
 Ha: Racerunners use their habitat non-randomly.

> Data used for this question: 
    > habitat transect data
        > percent occurrance of each type of substrate (habitat data) -- what is most common?
        > did cover effect it?
    > McElroy habitat data
        > speed vs. ground clutter crossed -- are speeds faster when more/less ground clutter is crossed?
           > lmer(speed ~ ground_clutter_crossed + (1 | ID))
        > bipedalism's effect on speed 
           > create new data set that only includes runs where bipedalism did or didn't occur
           > lmer(speed ~ bipedal + (1 | ID)
        > ground clutter crossed effect on bipedalism -- probably can't do because there aren't any definite 'Y's'
    > lab data  
        > how often bipedalism occurs on different substrates ?
        
      
        
        
        
        
        
```{r}
#Loading necessary packages
library(lme4)
library(lmerTest)
```

```{r}
#Importing datasets

mcelroy_habitat_data <- read.csv("~/College of Charleston/Thesis/Data/McElroy Habitat Data/McElroy_Habitat_Data.csv")   #full data set
mcelroy_habitat_data_omit <-  read.csv("~/College of Charleston/Thesis/Data/McElroy Habitat Data/McElroy_Habitat_Data_omit.csv")  #data set minus runs where bipedalism was unknown
field_transects <- read.csv("~/College of Charleston/Thesis/Data/Field_Transects_Substrates.csv")  #field transect data
field_transects_totals <-  read.csv("~/College of Charleston/Thesis/Data/Field_Transects_Substrate_Totals.csv")  #field transect substrate totals
field_transects_totals_condensed <- read.csv("~/College of Charleston/Thesis/Data/Field_Transects_Substrate_Totals_condensed.csv")
lab.data <- read.csv("~/College of Charleston/Thesis/Videos Digitized/Thesis_videos_digitized_condensed_final_10.15.20_removed_non-standstill_starts+NAs+negatives.csv")
```
> fix ID column - not headed properly


```{r}
field_transects_totals
#heavy mixed was most common substrate type in transect study
field_transects_totals_condensed
#mixed was most common grouped susbtrate type

#lizards therefore have to be at least somewhat good at running on heavy mixed because there's so much of it in their habitat
```


```{r}
speed_ground_cover <- lmer(speed ~ ground_clutter_crossed + (1 | ï..ID), data = mcelroy_habitat_data)
anova(speed_ground_cover)
summary(speed_ground_cover)
```
  
```{r}
speed_bipedal <- lmer(speed ~ bipedal + (1 | ï..ID), data = mcelroy_habitat_data_omit)
anova(speed_bipedal)
summary(speed_bipedal)
```

# what do these tell me? no p-value shown. They don't appear significant.




```{r}
#Bipedalism percentage per substrate and sex Histogram
ggplot(data = lab.data_omit, aes(as.numeric(bipedal_percent)))  + 
  geom_histogram(fill="dark green", bins = 20, binwidth = 10) +
  facet_wrap(~substrate_type + sex) +
  facet_grid(sex ~ substrate_type)

#mean, median & range of bipedal percentage separated into substrate type
summaryBy(bipedal_percent ~ substrate_type + sex, data = lab.data_omit,
          FUN = list(mean, median, min, max))

#males are bipedal more often on every substrate than females
```  
  
  
  
  