---
title: 'Question 3: Is the pattern observed in question #1 explained by question #2?'
author: "Kayla Pehl"
date: "10/16/2020"
output: html_document
---

Question 3: Is the pattern observed in question #1 explained by question #2?
 Ho: Racerunners run the same on all substrates.
 Ha: Racerunners avoid moving on suboptimal parts of habitat.
 
> Data used to answer this question:
    > lab experiment data
        > velocity vs. substrate type - lmer
        > acceleration vs. substrate type - lmer 
        > bipedalism vs. substrate type - lmer
        > bipedalism vs velocity - lmer
        > power vs. substrate type ? - lmer
        > time vs. substrate type ? - lmer
    > transect data
        > effect of available cover; where is cover most frequent (more cover = safer)
    > McElroy habitat data    
        > speed vs. ground clutter crossed -- are speeds faster when more/less ground clutter is crossed?
           > lmer(speed ~ ground_clutter_crossed + (1 | ID))
        > bipedalism's effect on speed 
           > create new data set that only includes runs where bipedalism did or didn't occur
           > lmer(speed ~ bipedal + (1 | ID)
        > video analysis
           

> Would've been really helpful to have had choice experiment data for this question, but since I don't have it, it may be a lot harder to answer this question. 


```{r}
#Distance vs. Number Clutter 
mcelroy_dist_clutter_lme3 <- lme(log10(NumberClutter+1) ~log10(Distance), random = ~1|ID, data = fieldspeed_final, na.action = na.omit)
anova(mcelroy_dist_clutter_lme3)  #p-value=0.0173  #significant
shapiro.test(residuals(mcelroy_dist_clutter_lme3))  #normal
hist(residuals(mcelroy_dist_clutter_lme3))
performance::r2(mcelroy_dist_clutter_lme3)  #good  #C=0.655  #M=0.118

ggplot(fieldspeed_final, aes(log10(Distance), log10(NumberClutter+1))) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "Distance Traveled (m)", y = "Number of Clutter Crossed") +
  theme(panel.background = element_blank()) +
  theme(axis.line.x = element_line(color = "black", size = 0.5),
        axis.line.y = element_line(color = "black", size = 0.5))
#trying to get dots to color-coordinate with ID, but when I put col=ID after Distance in aes, it messes up the lm line

#as distance increased, so did the amount of clutter crossed
```

```{r}
#Speed vs. Distance
mcelroy_dist_speed_lme <- lme(log10(Speed) ~ log10(Distance), random = ~1|ID, data = fieldspeed_final, na.action=na.omit)
anova(mcelroy_dist_speed_lme)  #p-value=0.0001  #significant
shapiro.test(residuals(mcelroy_dist_speed_lme))   #normal
hist(residuals(mcelroy_dist_speed_lme))
performance::r2(mcelroy_dist_speed_lme)  #good  #C=0.702  #M=0.449

ggplot(fieldspeed_final, aes(log10(Distance), log10(Speed))) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "Distance (m)", y = "Speed (m/s)") +
  theme(panel.background = element_blank()) +
  theme(axis.line.x = element_line(color = "black", size = 0.5),
        axis.line.y = element_line(color = "black", size = 0.5))

#as distance increased, so did speed
```