---
title: "Thesis_Stats_3.2.21"
author: "Kayla Pehl"
date: "3/2/2021"
output: html_document
---

#This file contains analyses copied over from Thesis_Stats_all for Question 2. Models altered based off of Eric's 3/2/21 comments (via email). 

```{r}
#Loading necessary packages
library(ggplot2)
library(lme4)
library(doBy)
library(car)
library(emmeans)
library(nlme)
library(lmtest)
```

```{r}
#Set working directory
setwd("~/College of Charleston/Thesis/R_Data")

#Importing data

RR_demo2 <- read.csv("RR_demo2.csv")
  #demographics including peak velocity and acceleration (all and divided by substrate) for each lizard
lab_data2 <- read.csv("lab_data2.csv")
  #includes peak velocity column and biped total column
lab_data2_omit <- lab_data2[!is.na(lab_data2$bipedal_percent), ]
  #includes peak velocity column and biped total column with non-bipedal rows omitted
#lab_data_omit <- lab_data[!is.na(lab_data$bipedal_percent), ]
  #lab_data without trials that were run quadrupedally with no bipedalism present
fieldspeed <-  read.csv("fieldspeed.csv")
  #final fieldspeed data; double checked with data sheets & it's correct 11.18.20
  #replacing all fieldspeed_final data and McElroy_habitat_data
fieldspeed_reduced <-  read.csv("fieldspeed_reduced.csv")
  #final fieldspeed data without 2 rows that did not have substrate type listed
fieldspeed_reduced_debris <- read.csv("~/College of Charleston/Thesis/R_Data/fieldspeed_reduced_debris.csv")
  #same as fieldspeed_reduced except 2 debris points were removed for ground clutter analysis b/c debris substrates are made up of obstacles
fieldspeed_jan <- read.csv("fieldspeed_jan.csv")
  #same as fieldspeed_reduced except rows without speed values were taken out; this enabled me to change the order of the boxplots
field_transect_data_totals <- read.csv("field_transect_data_totals.csv") 
  #field transect substrate totals
field_transect_data_totals_condensed <- read.csv("field_transect_data_totals_condensed.csv")
  #field transect substrate totals condensed into debris/mixed/vegetation/sand
field_transect_data_totals_heaviness <- read.csv("field_transect_data_totals_heaviness.csv")
  #field transect susbtrate totals broken into heavy/light/sand
field_transect_condensed <- read.csv("field_transect_condensed.csv")
  #field transect data broken into debris, mixed and sand to match lab data
fieldspeed_substrate_percent <- read.csv("fieldspeed_substrate_percent.csv")
  #field data with substrate percentages
```

QUESITON 2

1) Testing SVL, TL and SEX on Speed and Acc.  These are just simple linear models [example: lm(log10(peak_v)~SVL] because each lizard only has one observation for each variable on each substrate.   In fact, all of the mixed models are throwing an error because with only one observation per individual you cannot fit a mixed model!   I assume these are all not-signifcant (or maybe one is barely signicant).


```{r}
#Velocity 
vel_sex <- lm(log10(peak_v) ~ sex, data = RR_demo2)   
anova(vel_sex)   #p=0.1519

vel_svl <- lm(log10(peak_v) ~ svl, data = RR_demo2)   
anova(vel_svl)   #p=0.5978

vel_tl <- lm(log10(peak_v) ~ tail_length, data = RR_demo2)   
anova(vel_tl)   #p=0.1286

#neither sex, SVL, or tail length affect acceleration
```

```{r}
#Acceleration
acc_sex <- lm(log10(peak_acc) ~ sex, data = RR_demo2)   
anova(acc_sex)   #p=0.1003

acc_svl <- lm(log10(peak_acc) ~ svl, data = RR_demo2)   
anova(acc_svl)   #p=0.8229

acc_tl <- lm(log10(peak_acc) ~ tail_length, data = RR_demo2)   
anova(acc_tl)   #p=0.02722

#tail length affected acceleration but sex and SVL did not 
```


RR_demo3 <- read.csv("~/College of Charleston/Thesis/Videos Digitized/Thesis_videos_digitized_condensed_final_3.2.21_removed_non-standstill_starts_demographics3_aka_RR_demo3.csv")

biped_sex <- lm(peak_biped_percent ~ sex, RR_demo3)
anova(biped_sex)

biped_svl <- lm(peak_biped_percent ~ svl, RR_demo3)
anova(biped_svl)

biped_tl <- lm(peak_biped_percent ~ tail_length, RR_demo3)
anova(biped_tl)



2) Test the effects bipedal and substate on peak_v and acc and substrate on % biped. 

```{r}
#Velocity ~ Substrate 
lme.log.side.vmax.sub <- lme(log10(peak_v) ~ substrate_type*biped_combined, random = ~1|ID, data = lab_data2)
anova(lme.log.side.vmax.sub, type="marginal")  #biped & interaction significant

shapiro.test(residuals(lme.log.side.vmax.sub))  #normal
performance::r2(lme.log.side.vmax.sub)  #C=0.479, M=0.359
lme.log.side.vmax.sub.tukey <- emmeans(lme.log.side.vmax.sub, pairwise ~ substrate_type*biped_combined, adjust="tukey", data=lab_data2)
lme.log.side.vmax.sub.tukey  #some significant differences due to gait

ggplot(lab_data2, aes(y=peak_v, x=substrate_type, fill = biped_combined)) +
  geom_boxplot() + 
  scale_fill_discrete(labels = c("bipedal", "quadrupedal")) +
  labs(x = "Substrate", y = "Peak Velocity (m/s)") +
  scale_x_discrete(labels = c("Debris", "Sand", "Mixed")) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        panel.background = element_blank(), 
        axis.line = element_line(color = "black")) +
  theme(axis.line.x = element_line(color = "black", size = 0.5),
        axis.line.y = element_line(color = "black", size = 0.5),
        axis.text = element_text(face = "bold"),
        legend.title = element_blank())
```

```{r}
#Acceleration ~ Substrate
lme.log.side.amax.sub <- lme(log10(side_amax) ~ substrate_type*biped_combined, random = ~1|ID, data = lab_data2)
anova(lme.log.side.amax.sub, type="marginal")  #all insignificant

shapiro.test(residuals(lme.log.side.amax.sub))  #normal
performance::r2(lme.log.side.amax.sub)  #C=0.0.79, M=0.031
lme.log.side.amax.tukey2 <- emmeans(lme.log.side.amax.sub, pairwise ~ substrate_type*biped_combined, adjust="tukey", data=lab_data2)
lme.log.side.amax.tukey2   #no significant differences in any substrate's acceleration based on gait

ggplot(lab_data2, aes(y=side_amax, x=substrate_type, fill = biped_combined)) +
  geom_boxplot() + 
  scale_fill_discrete(labels = c("bipedal", "quadrupedal")) +
  labs(x = "Substrate", y = bquote('Peak Acceleration' ~ (m/s^2))) +
  scale_x_discrete(labels = c("Debris", "Sand", "Mixed")) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        panel.background = element_blank(), 
        axis.line = element_line(color = "black")) +
  theme(axis.line.x = element_line(color = "black", size = 0.5),
        axis.line.y = element_line(color = "black", size = 0.5),
        axis.text = element_text(face = "bold"),
        legend.title = element_blank())
```

```{r}
#Percent Bipedal ~ Substrate
lme.biped.sub <- lme(sqrt(bipedal_percent) ~ substrate_type, random = ~1|ID, data = lab_data2_omit)
anova(lme.biped.sub, type="marginal")  #p=0.0512 - substrate barely insignificant

shapiro.test(residuals(lme.biped.sub))  #normal
performance::r2(lme.biped.sub)  #C=0.417, M=0.050
lme.biped.sub.tukey <- emmeans(lme.biped.sub, pairwise ~ substrate_type, adjust="tukey", data=lab_data2_omit)
lme.biped.sub.tukey  #no significant differences in substrates

ggplot(lab_data2_omit, aes(y=bipedal_percent, x=substrate_type, fill = substrate_type)) +
  geom_boxplot() + 
  scale_fill_manual(values = c("lightskyblue2", "lightblue2", "steelblue")) +
  labs(x = "Substrate", y = "Percent Bipedal") +
  scale_x_discrete(labels = c("Debris", "Sand", "Mixed")) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        panel.background = element_blank(), 
        axis.line = element_line(color = "black")) +
  theme(axis.line.x = element_line(color = "black", size = 0.5),
        axis.line.y = element_line(color = "black", size = 0.5),
        axis.text = element_text(face = "bold"))
```



3) Field data:  i'd run these with lm.   I think we don't have enough data for the mixed model to work (too few repeated observations per indiv).   You will be treating each obs. as unique.  This is not ideal, because we know they are not (some obs. come from the same indiv)...but there are not enough data to really do the mixed model properly.    The analysis is what it is...if it's not significant then it's not

```{r}
#testing effect of sex on models

speed_sex <- lm(log10(Speed) ~ Sex, data = fieldspeed_jan, na.action=na.omit)
anova(speed_sex)  #p=0.692

clutter_sex <- lm(log10(NumberClutter+1) ~ Sex, data = fieldspeed, na.action = na.omit)
anova(clutter_sex)  #p=0.6028

#sex doesn't affect speed or clutter in field data
```

```{r}
#Speed ~ Substrate
mcelroy_speed_sub_lm <- lm(log10(Speed+1) ~ Substrate.Type, data = fieldspeed_jan, na.action=na.omit)
anova(mcelroy_speed_sub_lm)   #p=0.2832

mcelroy_speed_sub_lm.tukey <- emmeans(mcelroy_speed_sub_lm, pairwise ~ Substrate.Type, adjust="tukey", data=fieldspeed_jan)
mcelroy_speed_sub_lm.tukey

shapiro.test(residuals(mcelroy_speed_sub_lm))  #normal

ggplot(fieldspeed_jan, aes(x = reorder(Substrate.Type, Speed, FUN = median), y = Speed, fill = Substrate.Type)) + 
  geom_boxplot() +
  scale_fill_manual(values = c("lightskyblue2", "steelblue", "lightblue2")) +
  labs(x = "Substrate", y = "Speed (m/s)") +
  scale_x_discrete(labels = c("Debris", "Sand", "Mixed")) +
  theme(panel.background = element_blank()) +
  theme(axis.line.x = element_line(color = "black", size = 0.5),
        axis.line.y = element_line(color = "black", size = 0.5),
        axis.text = element_text(face = "bold"))
```

```{r}
#Speed ~ Number Clutter
mcelroy_speed_clutter_lm <- lm(log10(Speed+1) ~ log10(NumberClutter+1), data = fieldspeed_reduced_debris, na.action = na.omit)
anova(mcelroy_speed_clutter_lm)   #p=0.02004, significant

shapiro.test(residuals(mcelroy_speed_clutter_lm))   #normal

ggplot(fieldspeed_reduced_debris, aes(log(NumberClutter+1), log(Speed+1)))+
  geom_point(aes(color = Substrate.Type)) + 
  geom_smooth(method = 'lm') +
  labs(x = "log10(Ground Clutter Crossed)", y = "log10(Speed) (m/s)") +
  scale_color_discrete(name = "Substrate Type", labels = c("Mixed", "Sand")) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        panel.background = element_blank(), 
        axis.line = element_line(color = "black"),
        axis.text = element_text(face = "bold"))
```

```{r}
#testing non-logged data of Speed ~ Number Clutter
anova(lm(Speed ~ Substrate.Type, data = fieldspeed_jan, na.action=na.omit))  #p=0.3707
shapiro.test(residuals(lm(Speed ~ Substrate.Type, data = fieldspeed_jan, na.action=na.omit)))  #normal

anova(lm(Speed ~ NumberClutter, data = fieldspeed_reduced_debris, na.action = na.omit))   #p=0.03214, significant
shapiro.test(residuals(lm(Speed ~ NumberClutter, data = fieldspeed_reduced_debris, na.action = na.omit)))   #normal

#both non-logged models are worse than logged models
```



```{r}
library(multcompView)

lme.log.side.amax.sub <- lme(log10(side_amax) ~ substrate_type*biped_combined, random = ~1|ID, data = lab_data2)

ggplot(lab_data2, aes(y=side_amax, x=substrate_type, fill = biped_combined)) +
  geom_boxplot() + 
  scale_fill_discrete(labels = c("bipedal", "quadrupedal")) +
  labs(x = "Substrate", y = bquote('Peak Acceleration' ~ (m/s^2))) +
  scale_x_discrete(labels = c("Debris", "Sand", "Mixed")) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        panel.background = element_blank(), 
        axis.line = element_line(color = "black")) +
  theme(axis.line.x = element_line(color = "black", size = 0.5),
        axis.line.y = element_line(color = "black", size = 0.5),
        axis.text = element_text(face = "bold"),
        legend.title = element_blank())

Model4 <- aov(side_amax ~ substrate_type*biped_combined, data = lab_data2)
TUKEY <- TukeyHSD(Model4)
plot(TUKEY , las=1, col="brown")

generate_label_df <- function(TUKEY, variable){

    # Extract labels and factor levels from Tukey post-hoc 
    Tukey.levels <- TUKEY[[variable]][,6]
    Tukey.labels <- data.frame(multcompLetters(Tukey.levels)['Letters'])

    #I need to put the labels in the same order as in the boxplot :
    Tukey.labels$ChillTime=rownames(Tukey.labels)
    Tukey.labels=Tukey.labels[order(Tukey.labels$substrate_type) , ]
    return(Tukey.labels)
}

# Apply the function on my dataset
LABELS=generate_label_df(TUKEY , "substrate")

# A panel of colors to draw each group with the same color :
my_colors=c( rgb(143,199,74,maxColorValue = 255),rgb(242,104,34,maxColorValue = 255), rgb(111,145,202,maxColorValue = 255),rgb(254,188,18,maxColorValue = 255) , rgb(74,132,54,maxColorValue = 255),rgb(236,33,39,maxColorValue = 255),rgb(165,103,40,maxColorValue = 255))

# Draw the basic boxplot
a=boxplot(side_amax ~ substrate_type*biped_combined, data = lab_data2)#, col=my_colors[as.numeric(LABELS[,1])], ylab="value", main="")

# I want to write the letter over each box. Over is how high I want to write it.
over=0.1*max( a$stats[nrow(a$stats),] )

#Add the labels
text(c(1:nlevels(lme.log.side.amax.sub)) , a$stats[nrow(a$stats),]+over , LABELS[,1]  , col=my_colors[as.numeric(LABELS[,1])])

```


```{r}
vel <- structure(c(0.484L, .995L, .247L, .0001L, .0001L, .786L, .997L, .0002L, .0022L, .4899L, .0001L, .0001L, .0013L, .0106L, .982L), .Label = c("Debris", "Sand", "Mixed"), class = "factor"), gait = c(.321, .301, .316, .294, .225, .236)), .Names = c("substrate", "gait"), class = "data.frame")

vel$substrate <- factor(vel$substrate, levels = c("D", "S", "SD"))
```









