---
title: "Thesis_Stats_reduced"
author: "Kayla Pehl"
date: "2/21/2021"
output:
  word_document: default
  html_document: default
---

#This file contains models and plots from Thesis_Stats_all but only includes code that is being used in actual manuscript. All extra plots and code that are not currently in the manuscript are not in this file.

```{r}
#Loading necessary packages
library(ggplot2)
library(lme4)
library(doBy)
library(car)
library(emmeans)
library(nlme)
library(lmtest)
```

```{r}
#Set working directory
setwd("~/College of Charleston/Thesis/R_Data")

#Importing data

lab_data2 <- read.csv("lab_data2.csv")
  #includes peak velocity column and biped total column
lab_data2_omit <- lab_data2[!is.na(lab_data2$bipedal_percent), ]
  #includes peak velocity column and biped total column with non-bipedal rows omitted
#lab_data_omit <- lab_data[!is.na(lab_data$bipedal_percent), ]
  #lab_data without trials that were run quadrupedally with no bipedalism present
fieldspeed <-  read.csv("fieldspeed.csv")
  #final fieldspeed data; double checked with data sheets & it's correct 11.18.20
  #replacing all fieldspeed_final data and McElroy_habitat_data
fieldspeed_reduced <-  read.csv("fieldspeed_reduced.csv")
  #final fieldspeed data without 2 rows that did not have substrate type listed
fieldspeed_jan <- read.csv("fieldspeed_jan.csv")
  #same as fieldspeed_reduced except rows without speed values were taken out; this enabled me to change the order of the boxplots
field_transect_data_totals <- read.csv("field_transect_data_totals.csv") 
  #field transect substrate totals
field_transect_data_totals_condensed <- read.csv("field_transect_data_totals_condensed.csv")
  #field transect substrate totals condensed into debris/mixed/vegetation/sand
field_transect_data_totals_heaviness <- read.csv("field_transect_data_totals_heaviness.csv")
  #field transect susbtrate totals broken into heavy/light/sand
field_transect_condensed <- read.csv("field_transect_condensed.csv")
  #field transect data broken into debris, mixed and sand to match lab data
fieldspeed_substrate_percent <- read.csv("fieldspeed_substrate_percent.csv")
  #field data with substrate percentages
cover <- read.csv("cover.csv")
  #data of cover in the habitat assessment
field_obs_sub <- read.csv("field_obs_sub.csv")
  #field transect data - obstacles - broken into substrate type where the obstacles were found
field_obs_type <- read.csv("field_obs_type.csv")
  #field transect data - obstacles - broken into obstacle type
demo_debris <- read.csv("demo_debris.csv")
  #demographics and top performance on debris 
demo_mixed <- read.csv("demo_mixed.csv")
  #demographics and top performance on mixed
demo_sand <- read.csv("demo_sand.csv")
  #demographics and top performance on sand
demo_debris2 <- read.csv("demo_debris2.csv")
  #demographics and top performance on debris with lizards mizzing tails removed 
demo_mixed2 <- read.csv("demo_mixed2.csv")
  #demographics and top performance on mixed with lizards mizzing tails removed 
demo_sand2 <- read.csv("demo_sand2.csv")
  #demographics and top performance on sand with lizards mizzing tails removed 
RR_demo <- read.csv("RR_demo.csv")
  #demographics including peak velocity and acceleration (all and divided by substrate) for each lizard
```


Question 1: Do Racerunners use their habitat randomly or non-randomly?
 Ho: Racerunners use their habitat randomly.
 Ha: Racerunners use their habitat non-randomly.

```{r}
#bar plot of habitat assessment substrate totals condensed
ggplot(field_transect_condensed, aes(x=substrate, y=distance_m)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(x = "Substrate", y = "Distance (m)") +
  geom_text(aes(x=substrate, y = distance_m, label = percent_type, vjust = 1), size=4, 
            color = "black") +
  scale_x_discrete(labels = c("Debris", "Mixed", "Sand")) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        panel.background = element_blank(), 
        axis.line = element_line(color = "black")) +
  theme(axis.line.x = element_line(color = "black", size = 0.5),
        axis.line.y = element_line(color = "black", size = 0.5),
        axis.text = element_text(face = "bold"),
        plot.title = element_text(size = 15))
```

```{r}
#bar plot of field data substrate totals
ggplot(fieldspeed_substrate_percent, aes(x=Substrate.Type, y=Distance)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(x = "Substrate", y = "Distance Traveled (m)") +
  geom_text(aes(x=Substrate.Type, y = Distance, label = PercentSubstrate, vjust = 1), size=4, 
            color = "black") +
  scale_x_discrete(labels = c("Debris", "Mixed", "Sand")) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        panel.background = element_blank(), 
        axis.line = element_line(color = "black")) +
  theme(axis.line.x = element_line(color = "black", size = 0.5),
        axis.line.y = element_line(color = "black", size = 0.5),
        axis.text = element_text(face = "bold"))
```

```{r}
#bar plot of cover
ggplot(cover, aes(x=substrate, y=distance_m)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(x = "Substrate", y = "Distance Cover (m)") +
  geom_text(aes(x=substrate, y = distance_m, label = percent, vjust = 1), size=4, 
            color = "black") +
  scale_x_discrete(labels = c("Debris", "Mixed", "Sand")) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        panel.background = element_blank(), 
        axis.line = element_line(color = "black")) +
  theme(axis.line.x = element_line(color = "black", size = 0.5),
        axis.line.y = element_line(color = "black", size = 0.5),
        axis.text = element_text(face = "bold"))
```

```{r}
#bar plot of field obstacles per substrate
obs_sub <- field_obs_sub[c(2,3), ]   #removing debris from plot since I did not record obstacles there

ggplot(obs_sub, aes(x=substrate, y=number_obstacles)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(x = "Substrate", y = "Number of Obstacles per Substrate") +
  geom_text(aes(x=substrate, y = number_obstacles, label = percent, vjust = 1), size=4, 
            color = "black") +
  scale_x_discrete(labels = c("Mixed", "Sand")) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        panel.background = element_blank(), 
        axis.line = element_line(color = "black")) +
  theme(axis.line.x = element_line(color = "black", size = 0.5),
        axis.line.y = element_line(color = "black", size = 0.5),
        axis.text = element_text(face = "bold"),
        plot.title = element_text(size = 15))
```

```{r}
#Chi Square test to see if % habitat is related to % field data
env <- matrix(c(1899.5, 1.05, 1307.95, 16.94, 637.45, 34.928), ncol = 2, byrow = TRUE)
colnames(env) <- c("random", "used")
rownames(env) <- c("debris", "mixed", "sand")
env

chisq.test(env)  #p < 2.2e-16  #significant - distribution of proportions between the two groups is different
```


Question 2: How does "available" habitat affect Racerunner locomotion?
 Ho: Racerunners run the same on all substrates.
 Ha: Racerunners run fastest on sand compared to debris and mixed.

> Testing effect of sex, SVL, and tail length on velocity [Eric said I only need to test it on velocity]

```{r}
#Velocity vs Sex
lme.log.side.vmax.sex <- lme(log10(peak_v+1) ~ sex, random = ~1|ID, data = lab_data2)
anova(lme.log.side.vmax.sex)  #p-value =0.1399, insignificant
shapiro.test(residuals(lme.log.side.vmax.sex))  #normal
summary(lme.log.side.vmax.sex)
performance::r2(lme.log.side.vmax.sex)  #C=0.265, #M=0.032

#sex is insignificant for peak velocity
```

```{r}
#Peak Velocity (all) vs SVL 
lme.log.side.vmax.svl <- lme(log10(peak_v+1) ~ svl, random = ~1|ID, data = RR_demo)
anova(lme.log.side.vmax.svl)  #p-value =0.6383, insignificant
shapiro.test(residuals(lme.log.side.vmax.svl))  #normal
summary(lme.log.side.vmax.svl)
performance::r2(lme.log.side.vmax.svl)  #C=0.217, M=0.006

#SVL is insignificant on peak velocity
```

```{r}
#Peak Velocity (all) vs tail length 
lme.log.side.vmax.tail <- lme(log10(peak_v+1) ~ tail_length, random = ~1|ID, data = RR_demo)
anova(lme.log.side.vmax.tail)  #p-value =0.168, insignificant
shapiro.test(residuals(lme.log.side.vmax.tail))  #normal
summary(lme.log.side.vmax.tail)
performance::r2(lme.log.side.vmax.tail)  #C=0.214, M=0.048

#tail is insignificant on peak velocity
```

> Testing effect of substrate

```{r}
#Velocity vs Substrate Type - total
lme.log.side.vmax.sub2 <- lme(log10(peak_v+1) ~ substrate_type, random = ~1|ID, data = lab_data2)
anova(lme.log.side.vmax.sub2)  #p-value substrate=0.0001, significant  #sex insignificant
ggplot(lab_data2, aes(residuals(lme.log.side.vmax.sub2))) + 
  geom_histogram(bins = 11, color = "white") +
  labs(x = "Residuals of lme.log.side.vmax.sub2") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        panel.background = element_blank(), 
        axis.line = element_line(color = "black"))
shapiro.test(residuals(lme.log.side.vmax.sub2))  #normal
summary(lme.log.side.vmax.sub2)
performance::r2(lme.log.side.vmax.sub2)  #C=0.374 #M=0.128
lme.log.side.vmax.sub.tukey2 <- emmeans(lme.log.side.vmax.sub2, pairwise ~ substrate_type, adjust="tukey", data=lab_data2)
lme.log.side.vmax.sub.tukey2  #significant differences between D-S (p<0.0001) and D-SD (p=0.0405), but not S-SD (p=0.0661)

ggplot(lab_data2, aes(y=(peak_v), x=substrate_type, fill = substrate_type)) +
  geom_boxplot() + 
  scale_fill_manual(values = c("lightskyblue2", "lightblue2", "steelblue")) +
  labs(x = "Substrate", y = "Peak Velocity (m/s)") +
  scale_x_discrete(labels = c("Debris", "Sand", "Mixed")) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        panel.background = element_blank(), 
        axis.line = element_line(color = "black")) +
  theme(axis.line.x = element_line(color = "black", size = 0.5),
        axis.line.y = element_line(color = "black", size = 0.5),
        axis.text = element_text(face = "bold"))
```

```{r}
#Velocity vs Substrate + Bipedalism: Combined
lme.vmax.sub.biped3 <- lme(log10(peak_v+1) ~ substrate_type + biped_combined, random = ~1|ID, data = lab_data2)
anova(lme.vmax.sub.biped3)  #p_substrate < 0.0001,  p_biped_comb < 0.0001
ggplot(lab_data2, aes(residuals(lme.vmax.sub.biped3))) + 
  geom_histogram(bins = 11, color = "white") +
  labs(x = "Residuals of lme.vmax.sub.biped") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        panel.background = element_blank(), 
        axis.line = element_line(color = "black"))
shapiro.test(residuals(lme.vmax.sub.biped3))  #normal
summary(lme.vmax.sub.biped3)
performance::r2(lme.vmax.sub.biped3)  #C=0.466, M=0.335
lme.lme.vmax.sub.biped.tukey3 <- emmeans(lme.vmax.sub.biped3, pairwise ~ substrate_type, adjust="tukey", data=lab_data2)
lme.lme.vmax.sub.biped.tukey3  #D-S p=0.0006, D-SD p=0.0399, S-SD p=0.2526
  #bipedalism is significantly different between debris and sand and debris and mixed

ggplot(lab_data2, aes(y=peak_v, x=substrate_type, fill = biped_combined)) +
  geom_boxplot() + 
  scale_fill_discrete(labels = c("bipedal", "quadrupedal")) +
  labs(x = "Substrate", y = "Peak Velocity (m/s)") +
  scale_x_discrete(labels = c("Debris", "Sand", "Mixed")) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        panel.background = element_blank(), 
        axis.line = element_line(color = "black")) +
  theme(axis.line.x = element_line(color = "black", size = 0.5),
        axis.line.y = element_line(color = "black", size = 0.5),
        axis.text = element_text(face = "bold"),
        legend.title = element_blank())

#the bipedal/quadrupedal used in this model is separated into whether or not each lizard ran bipedal at any point during their trial. If bipedalism occurred at all, the entire run was marked bipedal; if the lizard did not appear to run bipedal at all, the run was marked quadrupedal.
```

```{r}
#Velocity vs Substrate + Bipedalism: Combined - Interaction Effect
lme.vmax.sub.biped4 <- lme(log10(peak_v+1) ~ substrate_type * biped_combined, random = ~1|ID, data = lab_data2)
anova(lme.vmax.sub.biped4)  #p_interaction = 0.0614
ggplot(lab_data2, aes(residuals(lme.vmax.sub.biped4))) + 
  geom_histogram(bins = 11, color = "white") +
  labs(x = "Residuals of lme.vmax.sub.biped") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        panel.background = element_blank(), 
        axis.line = element_line(color = "black"))
shapiro.test(residuals(lme.vmax.sub.biped4))  #normal
summary(lme.vmax.sub.biped4)
performance::r2(lme.vmax.sub.biped4)  #C=0.482, M=0.354
lme.lme.vmax.sub.biped.tukey4 <- emmeans(lme.vmax.sub.biped4, pairwise ~ substrate_type, adjust="tukey", data=lab_data2)
lme.lme.vmax.sub.biped.tukey4  #D-S p=0.0001, D-SD p=0.0081, S-SD p=0.3145
  #interaction between substrate*bipedalism is significantly different between debris-sand and debris-mixed

#interaction effect is overall not significant but it produces better p-value significance between D-S and D-SD than the non-interaction model
```

```{r}
#Acceleration vs Substrate Type - Side
lme.log.side.amax.sub <- lme(log10(side_amax+1) ~ substrate_type, random = ~1|ID, data = lab_data2)
anova(lme.log.side.amax.sub)  #p-value substrate & sex = not significant
ggplot(lab_data2, aes(residuals(lme.log.side.amax.sub))) + 
  geom_histogram(bins = 8, color = "white") +
  labs(x = "Residuals of lme.log.side.amax.sub") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        panel.background = element_blank(), 
        axis.line = element_line(color = "black"))
shapiro.test(residuals(lme.log.side.amax.sub))  #normal
summary(lme.log.side.amax.sub)
performance::r2(lme.log.side.amax.sub)   #C=0.047  #M=0.002
lme.log.side.amax.sub.tukey <- emmeans(lme.log.side.amax.sub, pairwise ~ substrate_type, adjust="tukey", data=lab_data2)
lme.log.side.amax.sub.tukey  #no significant differences in any substrate's acceleration

ggplot(lab_data2, aes(y=side_amax, x=substrate_type, fill = substrate_type)) +
  geom_boxplot() + 
  scale_fill_manual(values = c("lightskyblue2", "lightblue2", "steelblue")) +
  labs(x = "Substrate", y = bquote('Peak Acceleration' ~ (m/s^2))) +
  scale_x_discrete(labels = c("Debris", "Sand", "Mixed")) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        panel.background = element_blank(), 
        axis.line = element_line(color = "black")) +
  theme(axis.line.x = element_line(color = "black", size = 0.5),
        axis.line.y = element_line(color = "black", size = 0.5),
        axis.text = element_text(face = "bold"))

#low R2s make sense because the model shows that susbtrate type doesn't affect acceleration, which means that substrate type should not explain a high percentage of changes in the data

#substrate does not affect acceleration in side videos
```

```{r}
#Acceleration vs Substrate Type + Bipedalism Combined
lme.amax.sub.biped <- lme(log10(side_amax+1) ~ substrate_type + biped_combined, random = ~1|ID, data = lab_data2)
anova(lme.amax.sub.biped)  #p-value substrate & sex = not significant
ggplot(lab_data2, aes(residuals(lme.amax.sub.biped))) + 
  geom_histogram(bins = 8, color = "white") +
  labs(x = "Residuals of lme.amax.sub.biped") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        panel.background = element_blank(), 
        axis.line = element_line(color = "black"))
shapiro.test(residuals(lme.amax.sub.biped))  #normal
summary(lme.amax.sub.biped)
performance::r2(lme.amax.sub.biped)   #C=0.047  #M=0.002
lme.amax.sub.biped.tukey <- emmeans(lme.amax.sub.biped, pairwise ~ substrate_type, adjust="tukey", data=lab_data2)
lme.amax.sub.biped.tukey  #no significant differences in any substrate's acceleration

ggplot(lab_data2, aes(y=side_amax, x=substrate_type, fill = biped_combined)) +
  geom_boxplot() + 
  scale_fill_discrete(labels = c("bipedal", "quadrupedal")) +
  labs(x = "Substrate", y = bquote('Peak Acceleration' ~ (m/s^2))) +
  scale_x_discrete(labels = c("Debris", "Sand", "Mixed")) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        panel.background = element_blank(), 
        axis.line = element_line(color = "black")) +
  theme(axis.line.x = element_line(color = "black", size = 0.5),
        axis.line.y = element_line(color = "black", size = 0.5),
        axis.text = element_text(face = "bold"),
        legend.title = element_blank())
```

```{r}
#Acceleration vs Substrate Type + Bipedalism Combined - Interaction effect
lme.amax.sub.biped <- lme(log10(side_amax+1) ~ substrate_type * biped_combined, random = ~1|ID, data = lab_data2)
anova(lme.amax.sub.biped)  #p-value insignificant for all
ggplot(lab_data2, aes(residuals(lme.amax.sub.biped))) + 
  geom_histogram(bins = 8, color = "white") +
  labs(x = "Residuals of lme.amax.sub.biped") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        panel.background = element_blank(), 
        axis.line = element_line(color = "black"))
shapiro.test(residuals(lme.amax.sub.biped))  #normal
summary(lme.amax.sub.biped)
performance::r2(lme.amax.sub.biped)   #C=0.079  #M=0.030
lme.amax.sub.biped.tukey <- emmeans(lme.amax.sub.biped, pairwise ~ substrate_type, adjust="tukey", data=lab_data2)
lme.amax.sub.biped.tukey  #no significant differences in any substrate's acceleration

#interaction between substrate type and bipedalism was insignificant
```

```{r}
#Bipedalism vs Substrate Type
lme.biped.sub <- lme(sqrt(bipedal_percent) ~ substrate_type, random = ~1|ID, data = lab_data2_omit)
anova(lme.biped.sub)  #p-value substrate=0.0513  #barely not significant   #sex insignificant
ggplot(lab_data2_omit, aes(residuals(lme.biped.sub))) + 
  geom_histogram(bins = 7, color = "white") +
  labs(x = "Residuals of lme.biped.sub") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        panel.background = element_blank(), 
        axis.line = element_line(color = "black"))
shapiro.test(residuals(lme.biped.sub))   #normal
summary(lme.biped.sub)
performance::r2(lme.biped.sub)  #conditional: 0.426, marginal: 0.072  
lme.biped.sub.tukey <- emmeans(lme.biped.sub, pairwise ~ substrate_type, adjust="tukey", data=lab_data2_omit)
lme.biped.sub.tukey  #no significant differences in substrates

ggplot(lab_data2_omit, aes(y=bipedal_percent, x=substrate_type, fill = substrate_type)) +
  geom_boxplot() + 
  scale_fill_manual(values = c("lightskyblue2", "lightblue2", "steelblue")) +
  labs(x = "Substrate", y = "Percent Bipedal") +
  scale_x_discrete(labels = c("Debris", "Sand", "Mixed")) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        panel.background = element_blank(), 
        axis.line = element_line(color = "black")) +
  theme(axis.line.x = element_line(color = "black", size = 0.5),
        axis.line.y = element_line(color = "black", size = 0.5),
        axis.text = element_text(face = "bold"))

#substrate type does not affect bipedalism
```

```{r}
#Bipedalism's effect on peak velocity 
peakv.biped <- lme(log10(peak_v+1) ~ log10(bipedal_percent+1), random = ~1|ID, data = lab_data2_omit)
anova(peakv.biped)  #p-value substrate=0.0655, insignificant   #sex = insignificant
hist(residuals(peakv.biped))
shapiro.test(residuals(peakv.biped))  #normal
summary(peakv.biped)
performance::r2(peakv.biped)  #C=0.419  #M=0.075

ggplot(lab_data2_omit, aes(bipedal_percent, peak_v)) +
  geom_point() + 
  geom_smooth(method = 'lm') +
  labs(x = "Percent Bipedal", y = "Peak Velocity (m/s)") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        panel.background = element_blank(), 
        axis.line = element_line(color = "black"),
        axis.text = element_text(face = "bold"))

#bipedalism does not affect peak velocity
```

```{r}
#Bipedalism's effect on side acceleration
side.amax.biped <- lme(log10(side_amax+1) ~ log10(bipedal_percent+1) + sex, random = ~1|ID, data = lab_data2_omit)
anova(side.amax.biped)  #p-value substrate & sex = insignificant
hist(residuals(side.amax.biped))
shapiro.test(residuals(side.amax.biped))  #normal
summary(side.amax.biped)
performance::r2(side.amax.biped)  #C=0.130  #M=0.005

ggplot(lab_data2_omit, aes(bipedal_percent, side_amax)) +
  geom_point() + 
  geom_smooth(method = 'lm') +
  labs(x = "Percent Bipedal", y = bquote('Peak Acceleration' ~ (m/s^2))) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        panel.background = element_blank(), 
        axis.line = element_line(color = "black"),
        axis.text = element_text(face = "bold"))

#Bipedalism does not effect acceleration in side videos
```



Question 3: Is the pattern observed in question #1 explained by question #2?
 Ho: Racerunners run the same on all substrates.
 Ha: Racerunners avoid moving on suboptimal parts of habitat.


```{r}
#testing sex on field data: speed and ground clutter
mcelroy_speed_sex <- lme(log10(Speed) ~ Sex, random = ~1|ID, data = fieldspeed_jan, na.action=na.omit)
anova(mcelroy_speed_sex)  #p=0.6968

mcelroy_clutter_sex <- lme(log10(NumberClutter+1) ~ Sex, random = ~1|ID, data = fieldspeed, na.action = na.omit)
anova(mcelroy_clutter_sex)  #p=0.9755

#sex was insignficant for both speed and ground clutter in field data
```

```{r}
#Speed vs. Substrate Type 
mcelroy_speed_sub_lme3 <- lme(log10(Speed) ~ Substrate.Type, random = ~1|ID, data = fieldspeed_jan, na.action=na.omit)
anova(mcelroy_speed_sub_lme3)  #p-value substrate & sex = insignificant
summary(mcelroy_speed_sub_lme3)
mcelroy_speed_sub_lme_tukey3 <- emmeans(mcelroy_speed_sub_lme3, pairwise ~ Substrate.Type, adjust = "tukey", data = fieldspeed_jan)
mcelroy_speed_sub_lme_tukey3  #no substrates are signficant
hist(residuals(mcelroy_speed_sub_lme3))
shapiro.test(residuals(mcelroy_speed_sub_lme3))  #normal
performance::r2(mcelroy_speed_sub_lme3)

ggplot(fieldspeed_jan, aes(x = reorder(Substrate.Type, Speed, FUN = median), y = Speed, fill = Substrate.Type)) + 
  geom_boxplot() +
  scale_fill_manual(values = c("lightskyblue2", "steelblue", "lightblue2")) +
  labs(x = "Substrate", y = "Speed (m/s)") +
  scale_x_discrete(labels = c("Debris", "Sand", "Mixed")) +
  theme(panel.background = element_blank()) +
  theme(axis.line.x = element_line(color = "black", size = 0.5),
        axis.line.y = element_line(color = "black", size = 0.5),
        axis.text = element_text(face = "bold"))

#Substrate type doesn't appear to affect speed at all
```

```{r}
fieldspeed_reduced_debris <- read.csv("~/College of Charleston/Thesis/R_Data/fieldspeed_reduced_debris.csv")
#same as fieldspeed_reduced except 2 debris points were removed for ground clutter analysis because by definition, debris substrates were completely made up of obstacles

#Speed vs. Number Clutter
mcelroy_speed_clutter_lme_log4 <- lme(log(Speed+1) ~ log10(NumberClutter+1), random = ~1|ID, data = fieldspeed_reduced_debris, na.action = na.omit)
anova(mcelroy_speed_clutter_lme_log4)  #p-value=0.0276, significant
shapiro.test(residuals(mcelroy_speed_clutter_lme_log4))  #normal
hist(residuals(mcelroy_speed_clutter_lme_log4))
performance::r2(mcelroy_speed_clutter_lme_log4)  #ok  #C=0.145   #M=0.145
summary(mcelroy_speed_clutter_lme_log4)

#log-log plot
ggplot(fieldspeed_reduced_debris, aes(log(NumberClutter+1), log(Speed+1)))+
  geom_point(aes(color = fieldspeed_reduced_debris$`Substrate Type`)) + 
  geom_smooth(method = 'lm') +
  #labs(x = "Ground Clutter Crossed", y = "Speed (m/s)") +
  scale_color_discrete(name = "Substrate Type", labels = c("Mixed", "Sand")) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        panel.background = element_blank(), 
        axis.line = element_line(color = "black"),
        axis.text = element_text(face = "bold"))

#non-transformed plot
ggplot(fieldspeed_reduced_debris, aes(NumberClutter, Speed))+
  geom_point(aes(color = fieldspeed_reduced_debris$`Substrate Type`)) + 
  geom_smooth(method = 'lm') +
  labs(x = "Ground Clutter Crossed", y = "Speed (m/s)") +
  scale_color_discrete(name = "Substrate Type", labels = c("Mixed", "Sand")) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        panel.background = element_blank(), 
        axis.line = element_line(color = "black"),
        axis.text = element_text(face = "bold"))

#signficant relationship between speed and ground clutter crossed 
```






