---
title: "Thesis_Stats_all"
author: "Kayla Pehl"
date: "10/29/2020"
output: html_document
---

```{r}
#Loading necessary packages
library(ggplot2)
#library(rcompanion)
library(lme4)
#library(sjstats)
#library(doBy)
#library(car)
library(emmeans)
#library(multcompView)
library(nlme)
library(lmtest)
```

```{r}
#Importing data
lab_data <- read.csv("~/College of Charleston/Thesis/Videos Digitized/Thesis_videos_digitized_condensed_final_10.26.20_removed_non-standstill_starts+NAs+negatives+3lizards.csv")  
  #lab experiment data without no-standstill starts & 3 incomplete lizards
lab_data_omit <- lab_data[!is.na(lab_data$bipedal_percent), ]
  #lab_data without trials that were run quadrupedally with no bipedalism present
McElroy_habitat_data <- read.csv("~/College of Charleston/Thesis/Data/McElroy Habitat Data/McElroy_Habitat_Data2.csv")
  #McElroy & McBrayer habitat data from 2018-2019
field_transect_data_totals <- read.csv("~/College of Charleston/Thesis/Data/Field Transects/Field_Transects_Substrate_Totals.csv") 
  #field transect substrate totals
field_transect_data_totals_condensed <- read.csv("~/College of Charleston/Thesis/Data/Field Transects/Field_Transects_Substrate_Totals_condensed.csv")
  #field transect substrate totals condensed into debris/mixed/vegetation/sand
field_transect_data_totals_heaviness <- read.csv("~/College of Charleston/Thesis/Data/Field Transects/Field_Transects_Substrate_heaviness.csv")
  #field transect susbtrate totals broken into heavy/light/sand
```



Question 1: Do Racerunners use their habitat randomly or non-randomly?
 Ho: Racerunners use their habitat randomly.
 Ha: Racerunners use their habitat non-randomly.
 
```{r}
#Ground Clutter Crossed vs. Speed

log.speed.clutter <- lme(log10(speed) ~ log10(ground_clutter_crossed+1), random= ~1|ï..ID, data = McElroy_habitat_data, na.action=na.omit)
anova(log.speed.clutter)   #p-value = 0.0412  #significant
hist(residuals(log.speed.clutter))
shapiro.test(residuals(log.speed.clutter))  #normal  
performance::r2(log.speed.clutter)  #C=0.354 #M=0.124 

ggplot(McElroy_habitat_data, aes(log10(ground_clutter_crossed+1), log10(speed)) )+
  geom_point() + 
  geom_smooth(method = 'lm') +
  labs(x = "Ground Clutter Crossed", y = "Speed (m/s)") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(), 
        axis.line = element_line(color = "black"))
``` 

```{r}
#Field Transect Data
field_transect_data_totals  #heavy mixed most common
field_transect_data_totals_condensed  #mixed most common
field_transect_data_totals_heaviness  #heavy derbis/mixed/veg most common
```



Question 2: How does "available" habitat affect Racerunner locomotion?
 Ho: Racerunners run the same on all substrates.
 Ha: Racerunners run fastest on sand compared to debris and sand/debris.

```{r}
#Velocity vs Substrate Type - Side
lme.log.side.vmax.sub <- lme(log10(side_vmax+1) ~ substrate_type, random = ~1|ID, data = lab_data)
anova(lme.log.side.vmax.sub)  #p-value=0.0028  #significant
ggplot(lab_data, aes(residuals(lme.log.side.vmax.sub))) + 
  geom_histogram(bins = 11, color = "white") +
  labs(x = "Residuals of lme.log.side.vmax.sub") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(), 
        axis.line = element_line(color = "black"))
shapiro.test(residuals(lme.log.side.vmax.sub))  #normal (but just barely)
summary(lme.log.side.vmax.sub)
performance::r2(lme.log.side.vmax.sub)  #C=0.395 #M=0.056
lme.log.side.vmax.sub.tukey <- emmeans(lme.log.side.vmax.sub, pairwise ~ substrate_type, adjust="tukey", data=lab_data)
lme.log.side.vmax.sub.tukey  #significant differences between D-S but not D-SD or S-SD
ggplot(lab_data, aes(y=log10(side_vmax+1), x=substrate_type)) +
  geom_boxplot() + 
  labs(x = "Substrate", y = "Velocity (side camera) (m/s)") +
  scale_x_discrete(labels = c("Debris", "Sand", "Sand/Debris")) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(), 
        axis.line = element_line(color = "black"))
```

```{r}
#Velocity vs Substrate Type - Top

```

```{r}
#Acceleration vs Substrate Type - Side
lme.log.side.amax.sub <- lme(log10(side_amax+1) ~ substrate_type, random = ~1|ID, data = lab_data)
anova(lme.log.side.amax.sub)  #p-value=0.8411  #not significant
ggplot(lab_data, aes(residuals(lme.log.side.amax.sub))) + 
  geom_histogram(bins = 8, color = "white") +
  labs(x = "Residuals of lme.log.side.amax.sub") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(), 
        axis.line = element_line(color = "black"))
shapiro.test(residuals(lme.log.side.amax.sub))  #normal
summary(lme.log.side.amax.sub)
performance::r2(lme.log.side.amax.sub)   #C=0.047  #M=0.002
lme.log.side.amax.sub.tukey <- emmeans(lme.log.side.amax.sub, pairwise ~ substrate_type, adjust="tukey", data=lab_data)
lme.log.side.amax.sub.tukey  #no significant differences in any substrate's acceleration
ggplot(lab_data, aes(y=log10(side_amax+1), x=substrate_type)) +
  geom_boxplot() + 
  labs(x = "Substrate", y = "Acceleration (side camera) (m^2/s)") +
  scale_x_discrete(labels = c("Debris", "Sand", "Sand/Debris")) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(), 
        axis.line = element_line(color = "black"))

#low R2s make sense because the model shows that susbtrate type doesn't affect acceleration, which means that substrate type should not explain a high percentage of changes in the data
```

```{r}
#Bipedalism vs Substrate Type
lme.biped.sub <- lme(sqrt(bipedal_percent) ~ substrate_type, random = ~1|ID, data = lab_data_omit)
anova(lme.biped.sub)  #p-value=0.0512  #barely not significant
ggplot(lab_data_omit, aes(residuals(lme.biped.sub))) + 
  geom_histogram(bins = 7, color = "white") +
  labs(x = "Residuals of lme.biped.sub") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(), 
        axis.line = element_line(color = "black"))
shapiro.test(residuals(lme.biped.sub))   #normal
summary(lme.biped.sub)
performance::r2(lme.biped.sub)  #conditional: 0.417, marginal: 0.05  
lme.biped.sub.tukey <- emmeans(lme.biped.sub, pairwise ~ substrate_type, adjust="tukey", data=lab_data_omit)
lme.biped.sub.tukey  #no significant differences in substrates
ggplot(lab_data_omit, aes(y=sqrt(bipedal_percent), x=substrate_type)) +
  geom_boxplot() + 
  labs(x = "Substrate", y = "Percent Bipedal") +
  scale_x_discrete(labels = c("Debris", "Sand", "Sand/Debris")) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(), 
        axis.line = element_line(color = "black"))
```
#check back over - p-value barely insignificant


> Bipedalism on difference substrates

```{r}
#creating labels for histograms
sex.labs <- c("Female", "Male")
names(sex.labs) <- c("F", "M")
sub.labs <- c("Debris", "Sand", "Sand/Debris")
names(sub.labs) <- c("D", "S", "SD")

#Bipedalism per substrate and sex Histogram
ggplot(data = lab_data_omit, aes(as.numeric(bipedal_percent)))  + 
  geom_histogram(bins=10) +
  facet_wrap(~substrate_type + sex) +
  facet_grid(sex ~ substrate_type,
             labeller = labeller(sex=sex.labs, substrate_type=sub.labs)) +
  labs(x = "Percent Bipedal")


#mean, median & range of bipedal percentage separated into substrate type
summaryBy(bipedal_percent ~ substrate_type + sex, data = lab_data_omit,
          FUN = list(mean, median, min, max))
```


#FINISH DOING THIS HISTOGRAM AND THE ONE BELOW IT + BOXPLOT

```{r}
#Histogram of bipedal percent without NA's separated into substrate type and sex
ggplot(data = lab_data_omit, aes(as.numeric(bipedal_percent), col=substrate_type, fill=substrate_type, alpha=sex)) +
  geom_histogram(bins = 20, binwidth = 8) +
  facet_wrap(~substrate_type) +
  labs(x = "Percent Bipedal")
  

ggplot(data = lab_data_omit, aes(as.numeric(bipedal_percent), col=substrate_type, fill=substrate_type, alpha=sex)) +
  geom_histogram(bins = 20, binwidth = 8) +
  facet_wrap(~substrate_type) +
  labs(title = "Histogram of Bipedal Percent (without N/As) Separated by Substrate Type and Sex")  #original histogram

#Histogram of bipedal percent without N/A's broken into substrate type
ggplot(data = lab_data_omit, aes(as.numeric(bipedal_percent), col=substrate_type, fill=substrate_type)) +
  geom_histogram(col = "white") + 
  labs(title = "Histogram of Bipedal Percent (without N/As) Broken Down by Substrate Type")
```

```{r}
#Boxplot of bipedal percent without N/A's separated by substrate type and sex
ggplot(data = lab_data_omit, aes(substrate_type, as.numeric(bipedal_percent), fill = sex)) +
  geom_boxplot() +
  labs(x = "Substrate", y = "Percent Bipedal")
  scale_x_discrete(labels = c("Debris", "Sand", "Sand/Debris")) +
  theme(panel.background = element_blank())
```

